---
title: "Lecture 2 · Albert in a Linear World"
subtitle: "The simplest idea of change: a straight line plus shocks"
author: "Dynamic Systems Thinking for Psychologists"
date: "`r format(Sys.Date(), '%d %B %Y')`"
output:
  html_document:
    theme: flatly
    toc: true
    toc_float:
      collapsed: false
    toc_depth: 1
    df_print: kable
runtime: shiny
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE,
                      fig.align = "center")
library(shiny)
```

```{css, echo=FALSE}
body { font-size: 16px; }
h1 { margin-top: 1.6em; }
details {
  background: #f4f7fb;
  border-left: 4px solid #2a78d6;
  padding: 0.6em 1em;
  margin: 1em 0 1.6em 0;
  border-radius: 4px;
}
details > summary { cursor: pointer; font-weight: 600; color: #1c5cab; }
details[open] > summary { margin-bottom: 0.8em; }
.question {
  background: #fff8f2;
  border-left: 4px solid #eb6834;
  padding: 0.8em 1em;
  margin: 1.4em 0 0.4em 0;
  border-radius: 4px;
}
.question p:first-child { margin-top: 0; }
.question p:last-child  { margin-bottom: 0; }
.keybox {
  background: #eef4fc;
  border: 1px solid #b7d3f6;
  padding: 1em 1.2em;
  border-radius: 6px;
  margin: 1.4em 0;
}
.tool {
  border-top: 1px solid #d6d4cc;
  border-bottom: 1px solid #d6d4cc;
  padding: 0.6em 0;
  margin: 1.4em 0;
  color: #3d3c38;
}
.tool > p:first-child { margin-top: 0; }
.tool strong.tag {
  font-variant: small-caps; letter-spacing: 0.04em; color: #898781;
}
.report {
  background: #fbfaf6;
  border: 1px solid #e1e0d9;
  border-left: 4px solid #0b0b0b;
  font-family: Georgia, serif;
  padding: 1em 1.3em;
  margin: 1.4em 0;
}
.part {
  margin: 2.4em 0 0.4em 0;
  padding: 0.5em 0.9em;
  background: #0b0b0b;
  color: #fcfcfb;
  font-weight: 700;
  letter-spacing: 0.03em;
  border-radius: 4px;
}
.shiny-input-container { margin-bottom: 4px; }
```


```{r data, echo=TRUE}
# Albert's 21 weekly mood ratings (0 = worst imaginable, 100 = best imaginable).
# Week 0 is the intake session.
week <- 0:20
mood <- c(46, 35, 46, 46, 43, 30, 46, 45, 56, 53, 55,
          54, 59, 61, 44, 77, 73, 71, 64, 78, 71)

albert <- data.frame(week, mood)
n      <- nrow(albert)
```

```{r helpers, echo=FALSE}
# Plotting furniture and fast simulation. Nothing conceptual lives in here.

COL <- list(
  ink = "#0b0b0b", muted = "#898781", grid = "#e1e0d9",
  model = "#2a78d6", model_lt = "#9ec5f4", resid = "#eb6834",
  truth = "#e34948", dark = "#104281", up = "#2f9e6e"
)

alpha <- function(col, a)
  rgb(t(col2rgb(col)), alpha = a * 255, maxColorValue = 255)

base_par <- function(...)
  par(mar = c(4.4, 4.6, 3.2, 1.2), mgp = c(2.8, 0.7, 0), las = 1,
      col.axis = COL$muted, fg = COL$muted, tcl = -0.3,
      cex.lab = 1.1, cex.axis = 1.0, ...)

panel <- function(xlim, ylim, xlab = "Week of therapy", ylab = "Mood (0-100)",
                  main = "", sub = "", xat = NULL, yat = NULL) {
  plot(NA, xlim = xlim, ylim = ylim, xlab = xlab, ylab = ylab, axes = FALSE)
  if (is.null(xat)) xat <- pretty(xlim, 6)
  if (is.null(yat)) yat <- pretty(ylim, 6)
  abline(h = yat, v = xat, col = COL$grid, lwd = 1)
  axis(1, at = xat, col = COL$grid, col.ticks = COL$muted)
  axis(2, at = yat, col = COL$grid, col.ticks = COL$muted)
  if (nzchar(main))
    title(main = main, adj = 0, line = 1.8, cex.main = 1.3,
          col.main = COL$ink, font.main = 2)
  if (nzchar(sub))
    title(main = sub, adj = 0, line = 0.5, cex.main = 0.98,
          col.main = COL$muted, font.main = 1)
}

# Run code with a fixed seed WITHOUT disturbing the global random stream.
with_seed <- function(seed, code) {
  had <- exists(".Random.seed", envir = globalenv())
  if (had) old <- get(".Random.seed", envir = globalenv())
  on.exit(if (had) assign(".Random.seed", old, envir = globalenv())
          else rm(".Random.seed", envir = globalenv()))
  set.seed(seed)
  code
}

# One linear world, many Alberts: a (weeks x R) matrix of simulated moods.
linear_world <- function(R, b0, b1, sigma, weeks = 0:20)
  matrix(b0 + b1 * weeks, nrow = length(weeks), ncol = R) +
  matrix(rnorm(length(weeks) * R, 0, sigma), nrow = length(weeks), ncol = R)

# Simulate R replications of a study and fit each one, without calling lm().
sim_studies <- function(R, sigma, weeks = 0:20, b0 = b0_hat, b1 = b1_hat) {
  nn  <- length(weeks)
  xc  <- weeks - mean(weeks)
  Sxx <- sum(xc^2)
  Y   <- linear_world(R, b0, b1, sigma, weeks)
  b1h <- as.vector(crossprod(xc, Y)) / Sxx
  b0h <- colMeans(Y) - b1h * mean(weeks)
  rss <- colSums((Y - (outer(rep(1, nn), b0h) + outer(weeks, b1h)))^2)
  s   <- sqrt(rss / (nn - 2))
  se  <- s / sqrt(Sxx)
  tq  <- qt(0.975, nn - 2)
  data.frame(b0 = b0h, b1 = b1h, s = s, se = se,
             lo = b1h - tq * se, hi = b1h + tq * se)
}

# 95% joint confidence ellipse for (b0, b1).
conf_ellipse <- function(V, centre, df, level = 0.95, npts = 240) {
  th <- seq(0, 2 * pi, length.out = npts)
  ev <- eigen(V)
  A  <- ev$vectors %*% diag(sqrt(pmax(ev$values, 0)))
  r  <- sqrt(2 * qf(level, 2, df))
  t(centre + r * (A %*% rbind(cos(th), sin(th))))
}

# The three dials, reused by every Part 1 widget.
dials <- function(id, b0 = 40, b1 = 1.5, sigma = 8)
  list(
    sliderInput(paste0(id, "_b0"), "Starting level  (beta0):",
                min = 0, max = 100, value = b0, step = 1, width = "230px"),
    sliderInput(paste0(id, "_b1"), "Change per week  (beta1):",
                min = -3, max = 4, value = b1, step = 0.1, width = "230px"),
    sliderInput(paste0(id, "_sg"), "Size of weekly shocks  (sigma):",
                min = 0, max = 25, value = sigma, step = 0.5, width = "230px")
  )
```

```{r numbers, echo=FALSE}
# Every number quoted in the text is computed here, so the prose cannot drift
# away from the data.
fit    <- lm(mood ~ week, data = albert)
b0_hat <- unname(coef(fit)[1])
b1_hat <- unname(coef(fit)[2])
e      <- resid(fit)
s_hat  <- summary(fit)$sigma
se_b1  <- summary(fit)$coefficients["week", "Std. Error"]
t_b1   <- summary(fit)$coefficients["week", "t value"]
p_b1   <- summary(fit)$coefficients["week", "Pr(>|t|)"]
ci_b1  <- confint(fit)["week", ]
r2     <- summary(fit)$r.squared
Sxx    <- sum((week - mean(week))^2)
ybar   <- mean(albert$mood)
TSS    <- sum((albert$mood - ybar)^2)
RSS    <- sum(e^2)

d_alb     <- diff(mood)                           # Albert's week-to-week changes
worse_alb <- sum(d_alb < 0)
r_alb     <- cor(d_alb[-length(d_alb)], d_alb[-1])

fit_20   <- b0_hat + 20 * b1_hat
wk14_res <- unname(e[week == 14])
wk14_z   <- wk14_res / s_hat
p_low1   <- pnorm(-abs(wk14_z))                  # one week this low or lower
p_any21  <- 1 - (1 - p_low1)^n                   # at least one such week in 21
fit_no14 <- lm(mood ~ week, data = albert[week != 14, ])

pi21 <- predict(fit, data.frame(week = 21), interval = "prediction")
ci21 <- predict(fit, data.frame(week = 21), interval = "confidence")
hits_100 <- (100 - b0_hat) / b1_hat

f1    <- function(x, d = 1) formatC(x, format = "f", digits = d)
p_txt <- formatC(p_b1, format = "f", digits = 6)
```

# How to use this document

Open it in RStudio and click **Run Document** (or run
`rmarkdown::run("lecture01_linear_change.Rmd")`). It needs only the `shiny`
package.

- **Orange boxes** are questions. Answer them, in writing, *before* opening the
  blue box underneath. Committing to a wrong answer and finding out why is
  worth ten times more than reading the right one.
- **Sliders** are there to be moved. Most of the ideas here are much easier to
  feel than to read.
- **Grey-ruled boxes** name the statistical tool you have just used, so you
  can recognise it again outside of Albert's story.

---

# The course: what is making Albert's mood?

In 1920, John Watson and Rosalie Rayner taught an 11-month-old boy they called
**"Little Albert"** to fear a white rat, by striking a steel bar with a hammer
behind his head whenever he touched it. Within days Albert was afraid not
only of the rat but of a rabbit, a dog, a fur coat and a Santa Claus mask. He
left the hospital before anyone tried to undo the conditioning, and what
became of him is still debated.

In this course, Albert has grown up. Many years later he is in therapy for
his mood. At intake (week 0) and then once a week for twenty more weeks, his
therapist, Nora, asks him to mark his mood over the past week on a line from
*0 (worst imaginable)* to *100 (best imaginable)*. Those 21 numbers are our
data. (This grown-up Albert and his ratings are fictional.)

```{r fig-raw, echo=FALSE}
renderPlot({
  base_par()
  panel(c(-0.6, 20.6), c(20, 90), xat = seq(0, 20, 5),
        main = "Albert's mood over 21 weeks of therapy",
        sub  = "One self-rating per week, marked at the start of each session")
  points(albert$week, albert$mood, pch = 19, cex = 1.7, col = COL$ink)
  text(14, 44, "week 14", pos = 1, offset = 1.0, col = COL$muted, cex = 1.0)
}, height = 380)
```

Something produced these numbers: Albert's biology and history, his life
between sessions, the therapy, the act of marking a line on a sheet. We call
that the **data-generating mechanism**. We will never see it directly. What
we can do, and what this whole course does, is:

<div class="keybox">

1. **Propose** a mechanism, written as a model precise enough to *simulate*.
2. **Explore** the world that mechanism would produce: what kind of data does
   it make, and what can it never make?
3. **Confront** it with Albert: which version of it fits his data best, how
   uncertain is that fit, and where does his data look unlike anything the
   mechanism produces?

Each lecture proposes a richer mechanism than the last. Each one should get
us a little closer to whatever is really generating Albert's mood.

</div>

This first lecture starts with the simplest idea of change there is:
**therapy moves Albert along a straight line, and life adds a shock every week.** Part 1 explores what a
world like that looks like, before we touch Albert's data. Part 2 fits it to
Albert, and asks how much we can trust the fit.

::: {.question}
**Q1.** Before any model: how many mood points does Albert gain per week, on
average? Write your number down. You will need it in Part 2.
:::

---

<div class="part">Part 1 · The linear world</div>

# 1. A recipe for making an Albert

A data-generating mechanism is a **story about how each week's number comes
about**, told precisely enough that a computer could act it out. Here is the
first story, built in two steps.

### Step 1: what therapy does

Suppose therapy moves Albert's mood steadily. At intake his mood sits at some
level $\beta_0$, and each week it rises by the same amount $\beta_1$. After
$t$ weeks he is at

$$\beta_0 + \beta_1\, t$$

If that were the whole story, his chart would be a straight line of dots. Look
back at his real chart: it clearly isn't. Something else is happening every
week.

### Step 2: what life does

Go back to the list you made at the start: sleep, a row with a flatmate, the
weather, a good weekend, how he happened to read the scale that day. None of
that is in step 1. The linear story lumps all of it together into one term:
each week, life gives Albert a **shock** $\varepsilon_t$, a push up or down on
top of where therapy has brought him.

$$\text{Mood}_t \;=\; \underbrace{\beta_0 + \beta_1\, t}_{\text{what therapy does}}
\;+\; \underbrace{\varepsilon_t}_{\text{what life does}}$$

Saying *what kind* of shocks life deals is part of the story too. The linear
story makes three claims about them:

1. **Many small causes.** A week's shock is the sum of many small everyday
   influences, so it follows a bell curve centred on zero (Section 2 shows
   why).
2. **The same size every week.** How hard life hits Albert does not depend on
   the stage of therapy. One number, $\sigma$, describes how big shocks
   typically are.
3. **Forgotten by next week.** A shock does not carry over. Each week starts
   fresh.

### The story as code

Written week by week, the story is a loop:

```{r recipe, echo=TRUE}
b0 <- 40; b1 <- 1.5; sigma <- 8           # the three dials of this story
sim_mood <- numeric(21)
for (t in 0:20) {
  expected <- b0 + b1 * t                 # where therapy has brought him
  shock    <- rnorm(1, 0, sigma)          # what life did to him this week
  sim_mood[t + 1] <- expected + shock     # what he writes on the sheet
}
round(sim_mood)
```

Because the shocks are independent, R can do all 21 weeks in one line, and
statisticians write the same story in one line too:

$$\text{Mood}_t \;\sim\; \mathcal{N}\bigl(\,\beta_0 + \beta_1\, t,\;\; \sigma\,\bigr)
\qquad\Longleftrightarrow\qquad
\texttt{rnorm(21, b0 + b1 * week, sigma)}$$

Keep the loop in mind, though. Every later lecture tells a richer story by
editing a line in it; next lecture, for instance, lets this week's shock
depend on last week.

The story has three dials, and each one is a claim about Albert:

| Dial | Name | What it claims about Albert |
|---|---|---|
| $\beta_0$ | starting level | his *expected* mood at intake, before life's shocks |
| $\beta_1$ | change per week | how much therapy raises (or lowers) his expected mood each week, the same every week |
| $\sigma$ | shock size | how big life's weekly shocks typically are |

**Act the story out.** The blue line is what therapy does, his **expected
mood** $\beta_0 + \beta_1 t$. The orange strokes are what life did each week.
The dots are what Albert wrote on his sheet. Each press of the button lets
life deal a new set of shocks. The line is not itself an Albert: only a
story with $\sigma = 0$ puts him on it.

```{r fig-build, echo=FALSE}
do.call(inputPanel, c(dials("bd"), list(
  sliderInput("bd_weeks", "Weeks of therapy:", min = 5, max = 80, value = 20,
              step = 1, width = "230px"),
  checkboxInput("bd_shock", "Show this week's shocks", TRUE),
  checkboxInput("bd_six", "Act the story out six times", FALSE),
  actionButton("bd_go", "Let life deal new shocks")
)))

bd_sims <- reactive({
  input$bd_go
  wk <- 0:input$bd_weeks
  list(wk = wk, Y = linear_world(6, input$bd_b0, input$bd_b1, input$bd_sg, wk))
})

renderPlot({
  d <- bd_sims(); wk <- d$wk; Y <- d$Y
  mu <- input$bd_b0 + input$bd_b1 * wk
  yl <- range(c(0, 100, Y, mu))
  one <- function(j, main, sub) {
    panel(c(-0.5, max(wk) + 0.5), yl, xat = pretty(wk, 5), main = main, sub = sub)
    rect(-10, 100, 1e3, 1e3, col = alpha(COL$resid, 0.06), border = NA)
    rect(-10, -1e3, 1e3, 0, col = alpha(COL$resid, 0.06), border = NA)
    abline(h = c(0, 100), col = COL$resid, lty = 3)
    lines(wk, mu, col = COL$model, lwd = 3)
    if (isTRUE(input$bd_shock))
      segments(wk, mu, wk, Y[, j], col = COL$resid, lwd = if (length(wk) > 40) 1.2 else 2)
    points(wk, Y[, j], pch = 19, cex = if (length(wk) > 40) 0.9 else 1.4,
           col = COL$ink)
  }
  if (isTRUE(input$bd_six)) {
    par(mfrow = c(2, 3)); base_par(mar = c(3.6, 3.8, 2.6, 0.8))
    for (j in 1:6) one(j, sprintf("Run %d", j), "")
  } else {
    base_par()
    one(1, "One run of the story",
        sprintf("expected mood %.0f + %.1f t   +   shocks of size sigma = %.1f",
                input$bd_b0, input$bd_b1, input$bd_sg))
    legend("bottomright", bty = "n", lwd = c(3, 2, NA), pch = c(NA, NA, 19),
           col = c(COL$model, COL$resid, COL$ink), text.col = COL$ink, cex = 1.0,
           legend = c("what therapy does (expected mood)", "what life did (shock)",
                      "what Albert wrote down"))
  }
}, height = 440)
```

::: {.question}
**Q2.** Set $\sigma$ to 0. Describe the Albert this story produces. Does
anybody's mood work like that?

**Q3.** Set $\beta_1$ to 0: therapy does nothing at all. Keep $\sigma = 8$,
tick *act the story out six times*, and deal new shocks a few times. Do you
ever see an Albert who seems to be improving, or relapsing? What does that
tell you about reading a single person's chart?

**Q4.** Set $\beta_1 = 1.5$. With $\sigma = 3$, can you see what therapy is
doing? With $\sigma = 20$? What decides whether change is *visible*?

**Q5.** Set the therapy to 80 weeks. What does this story predict about
Albert's mood a year and a half in? Is that a problem with the story, or with
the question?
:::

<details>
<summary>Show the answers</summary>

**Q2.** Every week lands exactly on the line: Albert's mood is a clock. No
one's mood works like that. The shocks are where *everything the story does
not explain* goes. The linear story does not claim those influences are
unimportant or truly random; it just does not model them one by one, and
summarises their combined effect with a single dial, $\sigma$.

**Q3.** Yes, regularly. Twenty-one shocks contain runs, streaks and apparent
turning points, and the eye turns them into stories ("he was getting better
until week 12"). A story in which therapy does **nothing** still produces
charts that look like improvement or relapse. Keep this in mind for Part 2:
part of the job of statistics is to ask how easily the pattern we see could
come from shocks alone.

**Q4.** At $\sigma = 3$ the climb is unmistakable; at $\sigma = 20$ it
drowns. What matters is signal against shocks: the total change over therapy
($\beta_1 \times 20 = 30$ points) compared with the size of a typical shock
($\sigma$). The same real improvement can be obvious in one person and
invisible in another, simply because life knocks the second person around
more.

**Q5.** It predicts Albert keeps climbing at the same rate forever: past 100,
off the top of the scale. This story knows nothing about ceilings, plateaus or
a "normal" mood to settle back to. Over 20 weeks a straight line may be a fine
description; as a *mechanism* of change it cannot be right in the long run.
Later lectures tell stories that can level off.

</details>

::: {.tool}
<strong class="tag">The tool you just used</strong>

A **generative model**: $Y_t = \beta_0 + \beta_1 t + \varepsilon_t$ with
$\varepsilon_t \sim \mathcal{N}(0, \sigma)$ independently, written compactly
as $Y_t \sim \mathcal{N}(\beta_0 + \beta_1 t, \sigma)$. The line
$\beta_0 + \beta_1 t$ is the **expected value** $E[Y_t]$; the
$\varepsilon_t$ are called **errors**, **disturbances** or, in dynamic
models, **shocks** (or innovations).
:::

---

# 2. Life's shocks: why a bell curve?

The story claims that each week's shock follows a normal distribution. Why
should it? Nothing in Albert's life is "normally distributed". A row with his
flatmate either happens or it doesn't. A bad night's sleep costs a lot or
nothing. Most weeks bring no bad news; some bring a blow.

Start with a single influence, say sleep. In any one week it has one effect
on Albert's mood. But that effect **differs from week to week**: most weeks
sleep makes no difference, some weeks a bad night costs him a lot. The
pattern of how one influence's effect varies across weeks is *that
influence's distribution*, and it looks nothing like a bell.

A week's shock, though, is not one influence. It is the **sum of many**:
sleep, plus the flatmate, plus the weather, plus work, and so on, each
varying from week to week in its own way. Sums behave differently from their
parts. Below, each influence has a shape that is anything but a bell. Add
more of them together and watch what the *total* looks like, over 5000 of
Albert's weeks.

```{r fig-clt, echo=FALSE}
inputPanel(
  sliderInput("clt_k", "Small influences per week:", min = 1, max = 40,
              value = 1, step = 1, width = "260px"),
  selectInput("clt_shape", "Week to week, each influence is...",
              choices = c("a coin flip: a good or a bad day" = "coin",
                          "anything within a range" = "unif",
                          "mostly nothing, now and then a blow" = "skew"),
              selected = "skew", width = "300px")
)

clt_draw <- function(m, shape) switch(shape,
  coin = sample(c(-1, 1), m, replace = TRUE),
  unif = runif(m, -1, 1),
  skew = 1 - rexp(m))                       # mean 0, long tail of bad blows
clt_var <- list(coin = 1, unif = 1/3, skew = 1)   # variance of one influence

renderPlot({
  k <- input$clt_k; shape <- input$clt_shape
  one  <- clt_draw(20000, shape)
  tot  <- colSums(matrix(clt_draw(5000 * k, shape), nrow = k))
  tot  <- tot / sqrt(k * clt_var[[shape]]) * 8   # rescale: shocks of size sigma = 8

  par(mfrow = c(1, 2)); base_par()

  h1 <- hist(one, breaks = if (shape == "coin") c(-1.5, -0.5, 0.5, 1.5) else 40,
             plot = FALSE)
  plot(h1, freq = FALSE, col = alpha(COL$muted, 0.5), border = "white",
       main = "", xlab = "its effect on mood in a given week", ylab = "density", axes = FALSE)
  axis(1, col = COL$grid, col.ticks = COL$muted)
  axis(2, col = COL$grid, col.ticks = COL$muted)
  title("One influence, over 20000 weeks", adj = 0, line = 1.8, cex.main = 1.3, col.main = COL$ink)
  title("e.g. sleep: how much it moved his mood each week", adj = 0, line = 0.5, cex.main = 0.98,
        col.main = COL$muted, font.main = 1)

  br2 <- if (shape == "coin") seq(-k - 1, k + 1, by = 2) * 8 / sqrt(k) else 50
  h2 <- hist(tot, breaks = br2, plot = FALSE)    # coin sums sit on a lattice
  plot(NA, xlim = c(-32, 32), ylim = c(0, max(h2$density, dnorm(0, 0, 8)) * 1.25),
       axes = FALSE, xlab = "a week's shock (mood points)", ylab = "density")
  abline(h = pretty(c(0, max(h2$density)), 4), col = COL$grid)
  plot(h2, freq = FALSE, add = TRUE, col = alpha(COL$resid, 0.45), border = "white")
  curve(dnorm(x, 0, 8), add = TRUE, col = COL$model, lwd = 3)
  axis(1, col = COL$grid, col.ticks = COL$muted)
  axis(2, col = COL$grid, col.ticks = COL$muted)
  title(sprintf("The sum of %d influence%s", k, if (k == 1) "" else "s"),
        adj = 0, line = 1.8, cex.main = 1.3, col.main = COL$ink)
  title("5000 weeks; blue = normal curve",
        adj = 0, line = 0.5, cex.main = 0.98, col.main = COL$muted, font.main = 1)
}, height = 400)
```

::: {.question}
**Q6a.** Start with *mostly nothing, now and then a blow* and one influence
per week. Then add influences one at a time. Roughly how many does it take
before the total looks like the blue bell? Is it the same for the coin flip?

**Q6b.** What kind of week would break this argument, so that the shock is
*not* bell-shaped?
:::

<details>
<summary>Show the answers</summary>

**Q6a.** For the coin flip, a handful is enough; for the lopsided "blow now
and then" it takes more, a couple of dozen, before the long tail disappears.
This is the **central limit theorem**: add up enough small, independent
influences and the total is approximately normal, *whatever* the shape of the
individual influences. That is the rationale for normal shocks: not that any
single cause is normal, but that many small causes add up.

**Q6b.** A week dominated by *one* large event: a bereavement, a breakup,
losing a job. Then the shock is not the sum of many small things, and the
central limit theorem does not apply. The distribution of shocks then has
**heavier tails** than a bell: extreme weeks happen more often than the normal
curve predicts. The same happens if the small influences are not independent
(one bad night making the next more likely). Both are reasons a richer story
might replace the normal shock later on.

</details>

### One Albert's shocks

Because the story deals shocks from the same bell every week, we can look at
the bell through one Albert: here are the shocks from 1000 of his weeks, as
if he kept rating for twenty years. That makes it easy to ask how likely a
particular week is.

```{r fig-slice, echo=FALSE}
do.call(inputPanel, c(dials("sl"), list(
  sliderInput("sl_week", "In week:", min = 0, max = 20, value = 10,
              step = 1, width = "230px"),
  sliderInput("sl_cut", "How often is mood below...", min = 0, max = 100,
              value = 40, step = 1, width = "230px"),
  actionButton("sl_go", "Let life deal new shocks")
)))

sl_shocks <- reactive({ input$sl_go; rnorm(1000) })   # scaled by sigma below

renderPlot({
  b0 <- input$sl_b0; b1 <- input$sl_b1; sg <- max(input$sl_sg, 0.01)
  x0 <- input$sl_week; mu0 <- b0 + b1 * x0
  need <- input$sl_cut - mu0                # shock needed to fall below the cut
  sh <- sl_shocks() * sg
  y21 <- b0 + b1 * (0:20) + sh[1:21]
  yl <- range(c(0, 100, y21))

  par(mfrow = c(1, 2)); base_par()

  panel(c(-0.6, 20.6), yl, xat = seq(0, 20, 5),
        main = sprintf("Week %d: expected mood %.1f", x0, mu0),
        sub = sprintf("to fall below %d: a shock of %.1f or worse",
                      input$sl_cut, need))
  lines(0:20, b0 + b1 * (0:20), col = COL$model, lwd = 3)
  segments(0:20, b0 + b1 * (0:20), 0:20, y21, col = alpha(COL$resid, 0.6), lwd = 1.8)
  points(0:20, y21, pch = 19, cex = 1.2, col = alpha(COL$ink, 0.6))
  abline(h = input$sl_cut, col = COL$resid, lty = 2, lwd = 2)
  arrows(x0, mu0, x0, input$sl_cut, length = 0.08, col = COL$resid, lwd = 3)
  points(x0, mu0, pch = 19, cex = 1.8, col = COL$model)

  h <- hist(sh, breaks = 40, plot = FALSE)
  xl <- c(-4, 4) * sg
  plot(NA, xlim = xl, ylim = c(0, max(h$density, dnorm(0, 0, sg)) * 1.25), axes = FALSE,
       xlab = "a week's shock (mood points)", ylab = "density")
  abline(h = pretty(c(0, max(h$density)), 4), col = COL$grid)
  plot(h, freq = FALSE, add = TRUE, col = alpha(COL$resid, 0.45), border = "white")
  curve(dnorm(x, 0, sg), add = TRUE, col = COL$model, lwd = 3)
  abline(v = need, col = COL$resid, lty = 2, lwd = 2.5)
  axis(1, col = COL$grid, col.ticks = COL$muted)
  axis(2, col = COL$grid, col.ticks = COL$muted)
  title("1000 weeks of one Albert's shocks", adj = 0, line = 1.8,
        cex.main = 1.3, col.main = COL$ink)
  title(sprintf("that bad or worse: %.1f%% of weeks (theory %.1f%%)",
                100 * mean(sh < need), 100 * pnorm(need, 0, sg)),
        adj = 0, line = 0.5, cex.main = 0.98, col.main = COL$muted, font.main = 1)
}, height = 420)
```

::: {.question}
**Q6c.** In a story with $\beta_0 = 40$, $\beta_1 = 1.5$, $\sigma = 8$: how
often would Albert's mood in week 10 fall below 40, where he started?
Estimate first, then check.

**Q7.** Slide the week from 0 to 20. The shock needed changes, but does the
*bell* ever change? What claim about Albert is hidden in that?

**Q8.** About what share of weeks should land within ±2σ of the line? If you
watched Albert for 21 weeks, would you expect exactly that share?
:::

<details>
<summary>Show the answers</summary>

**Q6c.** Expected mood at week 10 is $40 + 1.5 \times 10 = 55$, so life would
have to push him down by 15 points, almost 2σ. Only about
`r f1(100 * pnorm(40, 55, 8))`% of weeks bring a shock that bad. After ten
weeks of steady improvement, a week as bad as intake is rare but not
impossible.

**Q7.** The bell never changes: the same shocks are dealt at intake and after
twenty weeks. So the linear story claims Albert is **exactly as unpredictable
at intake as after twenty weeks of therapy**. For someone recovering from
depression that is doubtful. People in an acute episode are often *more*
changeable than people who have stabilised, and instability is itself studied
as an early warning sign of relapse. A story in which the size of shocks
changes is a different, richer story.

**Q8.** About 95%. Over 21 weeks you would rarely see exactly that. The count
moves in steps of about 5 percentage points, and small samples scatter around
the long-run value. Expect anything from about 18 to 21 of 21.

</details>

::: {.tool}
<strong class="tag">The tools you just used</strong>

The **central limit theorem** (sums of many small independent influences are
approximately normal) as the rationale for **normal errors**; the
**conditional distribution** $Y_t \sim \mathcal{N}(\beta_0 + \beta_1 t,
\sigma)$, i.e. expected mood plus a normal shock; and two assumptions made
visible: **linearity** (the expected mood lies on a line) and
**homoscedasticity** (the same bell of shocks every week).
:::

---

# 3. How does a linear Albert change from one week to the next?

This course is about *dynamics*: how a person's state now relates to their
state a moment ago. So ask the linear story a dynamic question: **if therapy
is steadily improving Albert, how does he change from one week to the next?**

The answer follows from the story. The change from last week to this week is

$$\text{Mood}_t - \text{Mood}_{t-1} \;=\; \beta_1 \;+\; (\varepsilon_t - \varepsilon_{t-1})$$

the steady improvement $\beta_1$, plus this week's shock minus last week's.
Watch what that produces, first over 20 weeks and then over 1000 weeks of the
same Albert.

```{r fig-changes, echo=FALSE}
do.call(inputPanel, c(dials("ch", b0 = 40, b1 = 1.5, sigma = 8), list(
  actionButton("ch_go", "Let life deal new shocks")
)))

ch_sims <- reactive({
  input$ch_go
  y <- linear_world(1, input$ch_b0, input$ch_b1, max(input$ch_sg, 0.01),
                    weeks = 0:1000)[, 1]
  diff(y)                                   # 1000 week-to-week changes
})

renderPlot({
  D <- ch_sims()
  b1 <- input$ch_b1; sg <- max(input$ch_sg, 0.01)
  par(mfrow = c(1, 2)); base_par()

  ## left: the first 20 weeks
  d1 <- D[1:20]
  lim <- max(abs(d1), 5) * 1.15
  panel(c(0.4, 20.6), c(-lim, lim), xat = seq(0, 20, 5), ylab = "change since last week",
        main = "Albert, week by week",
        sub = sprintf("worse than the week before in %d of 20;  dashed = beta1", sum(d1 < 0)))
  abline(h = 0, col = COL$muted, lwd = 1.5)
  abline(h = b1, col = COL$model, lwd = 2.5, lty = 2)
  rect(1:20 - 0.35, 0, 1:20 + 0.35, d1, border = NA,
       col = ifelse(d1 < 0, COL$resid, COL$up))

  ## right: does a jump predict the next change?
  x <- D[-length(D)]; y <- D[-1]
  lim2 <- quantile(abs(c(x, y)), 0.995)
  plot(NA, xlim = c(-lim2, lim2), ylim = c(-lim2, lim2), axes = FALSE,
       xlab = "change this week", ylab = "change next week")
  abline(h = pretty(c(-lim2, lim2), 5), v = pretty(c(-lim2, lim2), 5), col = COL$grid)
  abline(h = 0, v = 0, col = COL$muted)
  points(x, y, pch = 19, cex = 0.6, col = alpha(COL$model, 0.35))
  abline(lm(y ~ x), col = COL$resid, lwd = 3)
  axis(1, col = COL$grid, col.ticks = COL$muted)
  axis(2, col = COL$grid, col.ticks = COL$muted)
  title("Jump now, change next?", adj = 0, line = 1.8,
        cex.main = 1.3, col.main = COL$ink)
  title(sprintf("1000 weeks: worse %.0f%% (theory %.0f%%)   r = %+.2f",
                100 * mean(D < 0), 100 * pnorm(0, b1, sg * sqrt(2)), cor(x, y)),
        adj = 0, line = 0.5, cex.main = 0.98, col.main = COL$muted, font.main = 1)
}, height = 420)
```

::: {.question}
**Q9.** Therapy improves Albert by a steady 1.5 points every week, and life
deals shocks of size $\sigma = 8$. In what share of weeks is he *worse* than
the week before? Guess before you look.

**Q10.** Look at the right panel. After a week in which Albert jumped up a
lot, what tends to happen the next week? Is there anything in the linear
story that pulls him back?

**Q11.** A clinician notices that whenever Albert has a very good week, he
"crashes" the week after, and concludes that he cannot sustain good moods.
What would you say?
:::

<details>
<summary>Show the answers</summary>

**Q9.** About `r round(100 * pnorm(0, 1.5, 8 * sqrt(2)))`%: nearly half.
A steadily improving person, with *nothing* going wrong in therapy, gets
worse from one week to the next almost as often as he gets better. The steady
improvement (1.5 points) is small compared with the scatter of a difference
between two shocks ($\sigma\sqrt{2} \approx 11$ points). Week-to-week
comparisons are a very noisy way to judge progress.

**Q10.** It tends to drop again: the correlation between successive changes
is about −0.5. Yet nothing in the linear story pulls Albert back; every
shock is forgotten by the following week. The pattern arises because a lucky
week's shock $\varepsilon_t$ appears in *both* changes: with a plus sign in
the jump up, and with a minus sign in the change after. Shocks with no memory
at all make a person *look* as if he bounces back.

**Q11.** That pattern is exactly what the linear story produces, and the
linear story contains no psychological process at all. Before we credit (or
blame) Albert with self-regulation, we need a story in which regulation
actually exists, and a way to tell it apart from this artefact. That is
precisely the kind of mechanism later lectures build: in a dynamic model,
last week's state really does pull on this week's.

</details>

::: {.tool}
<strong class="tag">The tool you just used</strong>

**Differencing** a time series, $\Delta Y_t = Y_t - Y_{t-1}$, and the
**independence** assumption: in the linear story
$\mathrm{Cor}(\varepsilon_t, \varepsilon_{t-1}) = 0$, which makes successive
differences correlate at exactly −0.5. Independence is the one assumption you
cannot see in a single week, and the one most likely to be false for repeated
ratings from one person.
:::

---

# 4. Could Albert live in a linear world?

Now bring in Albert, but still without fitting anything. The question is not
yet *which* linear world, only: **is Albert's data the kind of thing a linear
world makes at all?**

Two ways to ask it.

**Fingerprints.** Section 3 showed two fingerprints of the linear story: a
steadily improving Albert is worse than the week before in close to half of
weeks, and successive changes correlate at about −0.5. The real Albert was
worse than the week before in **`r worse_alb` of 20** weeks, and his
successive changes correlate at **r = `r f1(r_alb, 2)`**. Both are what a
linear world would produce.

**A lineup.** Below, Albert's real 21 weeks are hidden among eight runs of the
linear story, with dials that *you* set (rounded to whole numbers
and kept on the 0 to 100 scale, like real ratings). First tune the dials until
the fakes look like the real thing. Then give a neighbour the screen: can they
find Albert?

```{r fig-lineup, echo=FALSE}
do.call(inputPanel, c(dials("lu", b0 = 55, b1 = 0.5, sigma = 3), list(
  actionButton("lu_go", "New lineup"),
  checkboxInput("lu_show", "Reveal the real Albert", FALSE)
)))

lu_pos <- reactive({ input$lu_go; sample(9, 1) })
lu_Y   <- reactive({                     # redrawn by the dials or the button,
  input$lu_go                            # but NOT by the reveal checkbox
  Y <- linear_world(9, input$lu_b0, input$lu_b1, input$lu_sg)
  pmin(pmax(round(Y), 0), 100)
})

renderPlot({
  pos <- lu_pos()
  Y <- lu_Y()
  Y[, pos] <- albert$mood
  yl <- range(c(20, 90, Y))
  par(mfrow = c(3, 3)); base_par(mar = c(2.4, 3.2, 3.0, 0.6), mgp = c(2, 0.6, 0))
  for (j in 1:9) {
    real <- isTRUE(input$lu_show) && j == pos
    panel(c(-0.5, 20.5), yl, xlab = "", ylab = "", xat = seq(0, 20, 5),
          main = if (real) sprintf("%d  <- Albert", j) else as.character(j))
    points(week, Y[, j], pch = 19, cex = 1.1,
           col = if (real) COL$resid else COL$ink)
  }
}, height = 620)
```

::: {.question}
**Q12.** Which dial settings made the fakes hardest to tell from Albert? Write
them down next to your guess from Q1.

**Q13.** Suppose that, with good dial settings, nobody in the room can pick
out the real Albert. What have you shown, and what have you *not* shown?
:::

<details>
<summary>Show the answers</summary>

**Q12.** Most people end up somewhere near a starting level in the high 30s,
a weekly change around 1.5 to 2, and shocks of size around 7 or 8. Hold on to your
values: Part 2 does the same thing systematically, and you will be able to
see how close your eye got.

**Q13.** You have shown that a linear world *can* produce data like Albert's:
with only 21 weeks, his chart does not stand out from linear Alberts. You
have **not** shown that his mood *is* generated by a straight line plus
shocks. Many different mechanisms could produce 21 points like these,
including ones with memory, set points or tipping points, and short, noisy
series are bad at telling them apart. "The model fits" is the start of the
argument, not the end of it. That is why this course keeps proposing new
mechanisms and asking where each one fails.

</details>

<div class="keybox">

**What the linear world claims about how Albert changes**

1. **Change is steady.** The same $\beta_1$ points every week, forever, with no
   ceiling and no plateau.
2. **Change is driven by time, not by Albert.** Where he is this week has no
   influence on where he goes next.
3. **Fluctuations have no memory.** A bad week is a one-off shock that
   vanishes completely by the following week.
4. **Unpredictability never changes.** He is exactly as variable at intake as
   after twenty weeks of therapy.

Each of these is a hypothesis about the mechanism, not a fact about Albert.
Every later lecture relaxes at least one of them.

</div>

---

<div class="part">Part 2 · Which linear world is Albert's?</div>

Part 1 went from **dials to data**: pick $\beta_0, \beta_1, \sigma$, and the
recipe produces Alberts. Part 2 goes backwards, from **data to dials**:
given the one Albert we actually have, which dial settings are most
consistent with him? That backwards step is **estimation**. And because every
Albert is one noisy draw from his world, the dials we recover are uncertain.
The rest of this lecture is about *how* uncertain, and why.

<div class="keybox">

Keep two kinds of variability apart from the start:

**(A) How much Albert's weeks scatter around his expected mood.** This is
$\sigma$, a dial of the world. You met it in Part 1.

**(B) How much our *estimate* of his dials would scatter, if his 20 weeks
could be lived again.** This is the **standard error**. It is not a dial of
the world; it is a property of what one short, noisy series lets us learn.

**(B) is caused by (A).** The noisier Albert's weeks, the wobblier any line we
draw through them.

</div>

---

# 5. From data back to dials

### The flat world: "he's just having ups and downs"

Start with the most sceptical linear world: $\beta_1 = 0$. Albert does not
change; his weeks just scatter around a fixed level. If that were his world,
the best guess for any week would simply be his average mood.

```{r null-model, echo=TRUE}
ybar <- mean(albert$mood)                 # best guess in a flat world
TSS  <- sum((albert$mood - ybar)^2)       # how badly it misses, in total
c(average_mood = round(ybar, 1), total_squared_error = round(TSS))
```

```{r fig-null, echo=FALSE}
renderPlot({
  base_par()
  panel(c(-0.6, 20.6), c(20, 90), xat = seq(0, 20, 5),
        main = "The flat world: Albert is always the same",
        sub  = sprintf("Predict %.1f every week.  Total squared error = %.0f",
                       ybar, TSS))
  segments(albert$week, ybar, albert$week, albert$mood,
           col = COL$resid, lwd = 2.5)
  abline(h = ybar, col = COL$model, lwd = 3)
  points(albert$week, albert$mood, pch = 19, cex = 1.6, col = COL$ink)
  text(20.4, 27, "orange = how wrong the flat world is each week",
       adj = 1, col = COL$resid, cex = 1.0)
}, height = 380)
```

::: {.question}
**Q14.** In a flat world, the orange misses would be nothing but shocks. What would
they look like? What do they actually look like?
:::

<details>
<summary>Show the answer</summary>

In a flat world they would be scattered up and down in no particular order,
like the six flat-world runs you acted out in Q3. Instead they are almost all
*below* the average early on and almost all *above* it later. The flat world
is wrong in a way that depends on the week. A slope is exactly the device
for absorbing that pattern, and the habit of checking whether any pattern is
**left over** comes back in Section 7.

</details>

### Finding the best dials

For any setting of $\beta_0$ and $\beta_1$, measure each week's vertical miss,
square it, and add the squares up. The **least-squares** estimate is the
setting that makes this total as small as possible.

$$\text{squared error}(b_0, b_1) \;=\; \sum_t \bigl(\text{mood}_t - (b_0 + b_1\,t)\bigr)^2$$

This is not an arbitrary rule. In the normal linear world of Part 1, the dial
setting with the smallest squared error is also the one under which Albert's
actual data would have been **most probable** (the *maximum-likelihood*
estimate). Least squares asks: *which linear world makes the real Albert
least surprising?*

**Try to beat the computer.** Left: your line through Albert's weeks. Right:
a map of every possible setting of the first two dials, darker meaning better.

```{r fit-by-hand, echo=FALSE}
inputPanel(
  sliderInput("b0", "Starting level  (b0):",
              min = 25, max = 50, value = 45, step = 0.1, width = "260px"),
  sliderInput("b1", "Change per week  (b1):",
              min = 0, max = 3.5, value = 0.4, step = 0.01, width = "260px"),
  checkboxInput("show_best", "Show the least-squares line", FALSE)
)

ssr_fun  <- function(b0, b1) sum((albert$mood - (b0 + b1 * albert$week))^2)
b0_grid  <- seq(25, 50, length.out = 140)
b1_grid  <- seq(0,  3.5, length.out = 140)
SSRsurf  <- outer(b0_grid, b1_grid, Vectorize(ssr_fun))

renderPlot({
  my_ssr <- ssr_fun(input$b0, input$b1)
  par(mfrow = c(1, 2))
  base_par()

  panel(c(-0.6, 20.6), c(20, 95), xat = seq(0, 20, 5),
        main = "Your line",
        sub  = sprintf("squared error = %.0f    (best possible = %.0f)",
                       my_ssr, RSS))
  segments(albert$week, input$b0 + input$b1 * albert$week,
           albert$week, albert$mood, col = COL$resid, lwd = 2.5)
  if (isTRUE(input$show_best))
    abline(fit, col = COL$muted, lwd = 2.5, lty = 2)
  abline(a = input$b0, b = input$b1, col = COL$model, lwd = 3.5)
  points(albert$week, albert$mood, pch = 19, cex = 1.5, col = COL$ink)
  if (isTRUE(input$show_best))
    legend("topleft", bty = "n", lty = c(1, 2), lwd = c(3.5, 2.5),
           col = c(COL$model, COL$muted), text.col = COL$ink, cex = 1.0,
           legend = c("your line", "least squares"))

  brk  <- c(seq(RSS, quantile(SSRsurf, 0.7), length.out = 15),
            max(SSRsurf) + 1)
  cols <- colorRampPalette(c(COL$dark, COL$model, COL$model_lt,
                             "#cde2fb", "#fcfcfb"))(length(brk) - 1)
  image(b0_grid, b1_grid, SSRsurf, breaks = brk, col = cols, axes = FALSE,
        xlab = "starting level  (b0)", ylab = "change per week  (b1)")
  contour(b0_grid, b1_grid, SSRsurf, levels = brk[-length(brk)],
          add = TRUE, col = alpha("#fcfcfb", 0.5), drawlabels = FALSE)
  axis(1, col = COL$grid, col.ticks = COL$muted)
  axis(2, col = COL$grid, col.ticks = COL$muted)
  points(b0_hat, b1_hat, pch = 4, cex = 1.8, lwd = 3, col = COL$resid)
  points(input$b0, input$b1, pch = 21, cex = 2.2, lwd = 3,
         bg = alpha(COL$resid, 0.5), col = "#fcfcfb")
  title("Every possible line", adj = 0, line = 1.8, cex.main = 1.3,
        col.main = COL$ink)
  title("circle = you,  cross = the best line,  darker = better",
        adj = 0, line = 0.5, cex.main = 0.98, col.main = COL$muted,
        font.main = 1)
}, height = 420)
```

::: {.question}
**Q15.** Get your squared error as low as you can. How close are your dials
to the ones you picked by eye in the lineup (Q12)?

**Q16.** A colleague insists Albert improves by **2.11** points a week. Set
the starting level to **36.5** and the change to **2.11**, and note the error.
Keep the change at 2.11 and move the starting level until the error is as
small as you can make it. How much of the damage did you undo? What does that
say about the colleague's claim?
:::

<details>
<summary>Show the answer to Q16</summary>

At 36.5 and 2.11 the error is about `r round(ssr_fun(36.5, 2.11))`, against
a best possible `r round(RSS)`. Lowering the starting level to about
`r f1(ybar - 2.11 * mean(week))` brings it back to about
`r round(ssr_fun(ybar - 2.11 * mean(week), 2.11))`: you undo roughly
`r round(100 * (ssr_fun(36.5, 2.11) - ssr_fun(ybar - 2.11 * mean(week), 2.11)) / (ssr_fun(36.5, 2.11) - RSS))`%
of the damage.

So *"he started lower and is improving faster"* fits almost as well.
**Two wrong dials can make a nearly-right line.** On the map you moved a long
way from the cross but along the diagonal valley, so you stayed in the dark
region. The data do not pin down the starting level and the rate of change
separately: push one, and the other compensates. Section 11 shows where that
comes from, and what to do about it.

</details>

::: {.tool}
<strong class="tag">The tools you just used</strong>

The **null (intercept-only) model** and the **total sum of squares**
$\text{TSS} = \sum_t (y_t - \bar y)^2$; **ordinary least squares**, which
minimises the residual sum of squares $\text{RSS} = \sum_t (y_t - \hat y_t)^2$
and, under normal errors, coincides with **maximum likelihood**.
:::

---

# 6. The estimated dials

`lm()` finds the bottom of that valley with an exact formula.

```{r fit, echo=TRUE}
fit <- lm(mood ~ week, data = albert)
summary(fit)
```

Read the output as estimates of the three dials of Albert's linear world,
plus statements about how uncertain they are:

| In the output | Value | What it estimates, in Albert's terms |
|---|---|---|
| `(Intercept)` | `r f1(b0_hat, 2)` | $\beta_0$: his expected mood at intake, about `r round(b0_hat)` out of 100 |
| `week` | `r f1(b1_hat, 2)` | $\beta_1$: about `r f1(b1_hat)` points gained per week, roughly `r round(20 * b1_hat)` over 20 weeks |
| `Residual standard error` | `r f1(s_hat, 2)` | $\sigma$: a single week typically lands within about `r round(s_hat)` points of his expected mood. **Variability (A)** |
| `Std. Error` of `week` | `r f1(se_b1, 3)` | Not a dial: how much the estimated weekly gain would typically change if his 20 weeks were lived again. **Variability (B)** |
| `Multiple R-squared` | `r f1(r2, 2)` | Share of the variation in his ratings that the line accounts for: about `r round(100 * r2)`% |
| `t value` and p-value of `week` | `r f1(t_b1, 2)`, `r p_txt` | How surprising a slope this steep would be if his world were flat |

Look back at your guess from Q1 and your lineup dials from Q12.

### Could a flat world have produced this? Shuffling Albert's weeks

In a flat world ($\beta_1 = 0$) the *order* of Albert's weeks carries no
information: any ordering of his 21 ratings would be just as likely as the
real one. So shuffle them, fit a line to each shuffle, and see what slopes a
flat world tends to produce.

```{r fig-shuffle, echo=FALSE}
inputPanel(
  actionButton("sh_go", "Shuffle Albert's weeks 2000 more times")
)

sh_sims <- reactive({
  input$sh_go
  Y <- replicate(2000, sample(albert$mood))
  list(one = Y[, 1],
       b1  = as.vector(crossprod(week - mean(week), Y)) / Sxx)
})

renderPlot({
  d <- sh_sims()
  par(mfrow = c(1, 2)); base_par()

  panel(c(-0.6, 20.6), c(20, 90), xat = seq(0, 20, 5),
        main = "One shuffled Albert",
        sub  = sprintf("Same 21 ratings, random order.  slope = %+.2f",
                       d$b1[1]))
  abline(lm(d$one ~ week), col = COL$muted, lwd = 3, lty = 2)
  points(week, d$one, pch = 19, cex = 1.5, col = alpha(COL$ink, 0.7))

  hb <- hist(d$b1, breaks = 40, plot = FALSE)
  plot(NA, xlim = c(-2.3, 2.3), ylim = c(0, max(hb$density) * 1.3), axes = FALSE,
       xlab = "slope in a shuffled (flat) world", ylab = "density")
  abline(h = pretty(c(0, max(hb$density)), 4), col = COL$grid)
  plot(hb, freq = FALSE, add = TRUE, col = alpha(COL$muted, 0.5),
       border = "white")
  abline(v = b1_hat, col = COL$model, lwd = 3.5)
  text(b1_hat, max(hb$density) * 1.2, sprintf("Albert: %.2f", b1_hat),
       pos = 2, col = COL$model, font = 2, cex = 1.1)
  axis(1, col = COL$grid, col.ticks = COL$muted)
  axis(2, col = COL$grid, col.ticks = COL$muted)
  title("2000 shuffled Alberts", adj = 0, line = 1.8, cex.main = 1.3,
        col.main = COL$ink)
  title(sprintf("steepest: %.2f     as steep as Albert's: %d",
                max(abs(d$b1)), sum(abs(d$b1) >= b1_hat)),
        adj = 0, line = 0.5, cex.main = 0.98, col.main = COL$muted,
        font.main = 1)
}, height = 400)
```

::: {.question}
**Q17.** How often does a shuffled, flat-world Albert produce a slope as
steep as the real one? What does the p-value in the `lm()` output therefore
mean?

**Q18.** Does a tiny p-value show that Albert's world is *linear*?
:::

<details>
<summary>Show the answers</summary>

**Q17.** Essentially never. Shuffled slopes pile up around zero and mostly
stay within about ±1; Albert's `r f1(b1_hat, 2)` is far out in the tail. That
is what the p-value says: *if Albert's world were flat, data with a trend
this strong would almost never arise.* `lm()` computes it from the
$t$ distribution rather than by shuffling ($p \approx$ `r p_txt`), but the
question is the same. It is **not** the probability that the flat world is
true, nor the probability that therapy works.

**Q18.** No. It rules out one very specific world (flat, with independent
shocks) as a good account of Albert. It says nothing about whether a
straight line is the *right* shape, and nothing about whether the shocks are
really memoryless. Many non-linear mechanisms would also produce a
"significant" slope here.

</details>

::: {.tool}
<strong class="tag">The tools you just used</strong>

The **coefficient estimates** $\hat\beta_0, \hat\beta_1$, the **residual
standard error** $\hat\sigma$, the **standard error**
$\mathrm{SE}(\hat\beta_1)$, $R^2 = 1 - \text{RSS}/\text{TSS}$, and the
**p-value** for $H_0\!: \beta_1 = 0$, here illustrated by a **permutation
test**. With one predictor the F-test in the output is the same test:
$F = t^2$ (`r f1(t_b1, 2)`² = `r f1(t_b1^2, 1)`).
:::

---

# 7. The shocks we estimated: residuals, sigma, and week 14

Each week's miss from the fitted line, its **residual**, is our best guess at
that week's shock $\varepsilon_t$. If Albert lives in a linear world, the 21
residuals should look like 21 independent draws from one normal distribution
centred at zero, and their spread estimates $\sigma$:

$$\hat\sigma \;=\; \sqrt{\frac{\sum_t e_t^2}{n - 2}}$$

(dividing by $n - 2$ because two dials, $\beta_0$ and $\beta_1$, were already
estimated from these same 21 weeks).

```{r sigma-hat, echo=TRUE}
e <- resid(fit)
c(sum_of_squared_misses = round(sum(e^2), 1),
  divided_by_n_minus_2  = round(sum(e^2) / (n - 2), 1),
  sigma_hat             = round(sqrt(sum(e^2) / (n - 2)), 2))
```

```{r fig-resid, echo=FALSE}
renderPlot({
  par(mfrow = c(1, 3)); base_par()

  panel(c(-0.6, 20.6), c(-24, 24), ylab = "miss from line (points)",
        xat = seq(0, 20, 5),
        main = "Residuals over time", sub = "Any pattern left?")
  abline(h = 0, col = COL$model, lwd = 2.5)
  abline(h = c(-2, 2) * s_hat, col = COL$muted, lty = 3, lwd = 1.5)
  segments(albert$week, 0, albert$week, e, col = COL$resid, lwd = 2.5)
  points(albert$week, e, pch = 19, cex = 1.4, col = COL$ink)
  text(14, wk14_res, "week 14", pos = 4, col = COL$resid, font = 2, cex = 1.0)

  h <- hist(e, breaks = seq(-25, 25, 5), plot = FALSE)
  plot(NA, xlim = c(-25, 25),
       ylim = c(0, max(h$density, dnorm(0, 0, s_hat)) * 1.25), axes = FALSE,
       xlab = "residual", ylab = "density")
  abline(h = pretty(c(0, max(h$density)), 4), col = COL$grid)
  plot(h, freq = FALSE, add = TRUE, col = alpha(COL$model_lt, 0.8),
       border = "white")
  curve(dnorm(x, 0, s_hat), add = TRUE, col = COL$model, lwd = 3)
  axis(1, col = COL$grid, col.ticks = COL$muted)
  axis(2, col = COL$grid, col.ticks = COL$muted)
  title("Their distribution", adj = 0, line = 1.8, cex.main = 1.25,
        col.main = COL$ink)
  title(sprintf("Normal(0, %.1f) overlaid", s_hat), adj = 0, line = 0.5,
        cex.main = 0.95, col.main = COL$muted, font.main = 1)

  qq <- qqnorm(e, plot.it = FALSE)
  plot(NA, xlim = range(qq$x) * 1.15, ylim = range(qq$y) * 1.15, axes = FALSE,
       xlab = "expected if normal", ylab = "Albert's residual")
  abline(h = pretty(qq$y, 5), v = pretty(qq$x, 5), col = COL$grid)
  qqline(e, col = COL$model, lwd = 2.5)
  points(qq$x, qq$y, pch = 19, cex = 1.4, col = COL$ink)
  axis(1, col = COL$grid, col.ticks = COL$muted)
  axis(2, col = COL$grid, col.ticks = COL$muted)
  title("Normal Q-Q", adj = 0, line = 1.8, cex.main = 1.25, col.main = COL$ink)
  title("On the line = bell-shaped", adj = 0, line = 0.5, cex.main = 0.95,
        col.main = COL$muted, font.main = 1)
}, height = 340)
```

::: {.question}
**Q19.** In week 14 Albert rated his mood `r f1(-wk14_res)` points below the
line, `r f1(abs(wk14_z))` times $\hat\sigma$. Was it a relapse? Answer as the
linear world would: how likely is one week this bad, and how likely is it
that *at least one* of 21 weeks is this bad?

**Q20.** Drop week 14 and refit:
`lm(mood ~ week, data = albert[albert$week != 14, ])`. How much do the
estimated weekly gain and its standard error change? Should you drop it?
:::

<details>
<summary>Show the answers</summary>

**Q19.** In a linear world with $\sigma \approx$ `r f1(s_hat)`, a single week
this low has a probability of about `r f1(100 * p_low1)`%. But week 14 caught
our eye *because* it was the worst of 21. The fair question is how often 21
independent weeks contain at least one this bad:
$1 - (1 - `r signif(p_low1, 2)`)^{21} \approx$ `r round(100 * p_any21)`%, about
one course of therapy in `r round(1 / p_any21)`. Within the linear world,
week 14 needs no special explanation: the weeks after it are back on the
line. Clinically it is still worth asking Albert what happened. And a
mechanism with *memory* would read week 14 differently: as a shock whose
effect might linger. That is a question for a later lecture.

**Q20.** Without week 14 the weekly gain moves from `r f1(b1_hat, 2)` to
`r f1(coef(fit_no14)[2], 2)` and its standard error from `r f1(se_b1, 3)` to
`r f1(summary(fit_no14)$coefficients[2, 2], 3)`. The story is the same. But
a data point is only "bad" if it is *wrong* (a typo, another client's sheet).
Week 14 is a real week; dropping it would make Albert look steadier and
faster-improving than he is. Keep it.

</details>

::: {.tool}
<strong class="tag">The tools you just used</strong>

**Residuals** $e_t = y_t - \hat y_t$ as estimates of $\varepsilon_t$; the
**residual standard deviation** $\hat\sigma$ with $n - 2$ **degrees of
freedom**; **residual diagnostics**; **standardised residuals**
$e_t / \hat\sigma$; and the **multiple-comparisons** problem.
:::

---

# 8. If Albert lived these 20 weeks again

We now have estimated dials. But remember how the story works: every week
life deals Albert a fresh shock. The 21 shocks he actually got are only one
of the sets life could have dealt. Had his 20 weeks gone differently in all
the small everyday ways, the same therapy would have produced a different
chart, and we would have estimated different dials.

In Part 1 we acted the story out to see **what it produces**. Now we act it
out again for a new reason: to see **how much our estimate depends on which
shocks Albert happened to get**. We cannot rerun Albert's life, but we can
rerun the story, with its dials set to Albert's estimates, thousands of
times. Each run deals a fresh set of 21 shocks; fitting a line to each run
shows how much the estimate would wobble.

**Drag the sliders and watch what makes the estimates agree or disagree.**

```{r fig-universes, echo=FALSE}
inputPanel(
  sliderInput("pu_R", "How many reruns?",
              min = 20, max = 2000, value = 200, step = 20, width = "240px"),
  sliderInput("pu_sigma", "Size of weekly shocks (sigma):",
              min = 2, max = 20, value = round(s_hat), step = 1, width = "240px"),
  sliderInput("pu_n", "How many ratings:",
              min = 5, max = 60, value = 21, step = 1, width = "240px"),
  radioButtons("pu_span", "Extra ratings come from...",
               choices = c("a longer therapy (one per week)" = "longer",
                           "rating more often within the same 20 weeks" = "denser"),
               selected = "longer", width = "260px"),
  actionButton("pu_go", "Rerun again")
)

pu_sims <- reactive({
  input$pu_go
  wk <- if (input$pu_span == "longer") 0:(input$pu_n - 1)
        else seq(0, 20, length.out = input$pu_n)
  list(weeks = wk,
       sims  = sim_studies(input$pu_R, input$pu_sigma, weeks = wk))
})

renderPlot({
  d  <- pu_sims(); sims <- d$sims; wk <- d$weeks
  se_theory <- input$pu_sigma / sqrt(sum((wk - mean(wk))^2))

  par(mfrow = c(1, 2)); base_par()

  ends <- c(sims$b0, sims$b0 + sims$b1 * max(wk))
  yl   <- range(ends) + c(-1, 1) * 0.08 * diff(range(ends))
  panel(c(-0.6, max(wk) + 0.6), yl, xat = pretty(wk, 5),
        main = sprintf("%d reruns of Albert's therapy", input$pu_R),
        sub  = "Each faint line is the line fitted to one rerun")
  for (i in seq_len(min(nrow(sims), 300)))
    abline(a = sims$b0[i], b = sims$b1[i],
           col = alpha(COL$model, 0.08), lwd = 1.5)
  abline(a = b0_hat, b = b1_hat, col = COL$truth, lwd = 3.5)
  legend("topleft", bty = "n", lty = 1, lwd = 3.5, col = COL$truth,
         text.col = COL$ink, cex = 1.0,
         legend = "world's dials = Albert's estimates")

  hb <- hist(sims$b1, breaks = 40, plot = FALSE)
  xr <- b1_hat + c(-4, 4) * max(se_theory, 0.05)
  plot(NA, xlim = xr,
       ylim = c(0, max(hb$density, dnorm(0, 0, se_theory)) * 1.3),
       axes = FALSE, xlab = "estimated change per week", ylab = "density")
  abline(h = pretty(c(0, max(hb$density)), 4), col = COL$grid)
  plot(hb, freq = FALSE, add = TRUE, col = alpha(COL$model_lt, 0.85),
       border = "white")
  curve(dnorm(x, b1_hat, se_theory), add = TRUE, col = COL$model, lwd = 3)
  axis(1, col = COL$grid, col.ticks = COL$muted)
  axis(2, col = COL$grid, col.ticks = COL$muted)
  abline(v = b1_hat, col = COL$truth, lwd = 3)
  title("How much the estimates disagree", adj = 0, line = 1.8,
        cex.main = 1.3, col.main = COL$ink)
  title(sprintf("spread %.3f     theory sigma/sqrt(Sxx) %.3f",
                sd(sims$b1), se_theory),
        adj = 0, line = 0.5, cex.main = 0.98, col.main = COL$muted,
        font.main = 1)
}, height = 420)
```

<div class="keybox">

**That histogram is what a standard error is trying to describe.** We only
ever get one draw from it, Albert's actual 21 weeks, and from that single
draw `lm()` estimates how wide the histogram is:

$$\mathrm{SE}(\hat\beta_1) \;=\; \frac{\sigma}{\sqrt{S_{xx}}},
\qquad S_{xx} = \sum_t (t - \bar t)^2$$

Notice what is *not* in the formula: $\beta_0$ and $\beta_1$. The spread
depends only on the size of the shocks and on when the ratings were taken, which is why it
does not matter much that we set this world's dials to our estimates rather
than to Albert's unknown true ones.

</div>

::: {.question}
**Q21.** Double the size of the shocks from 8 to 16. What happens to the spread of the
estimates, and by what factor?

**Q22.** Put $\sigma$ back to 8. You could collect 60 ratings instead of 21 in
two ways: (a) see Albert weekly for 60 weeks; (b) keep therapy at 20 weeks but
have him rate his mood about three times a week. Try both. Which steadies the
estimate more, and why?

**Q23.** The fan of lines is pinched in the middle and splayed at the ends.
Why? What does that mean for how well we know Albert's mood *at intake* versus
*at week 10*?
:::

<details>
<summary>Show the answers</summary>

**Q21.** It exactly doubles: $\sigma$ is in the numerator. This is the causal
link between (A) and (B). A person whose weeks swing more gives a less
certain answer to "is he improving?", with the same real improvement.

**Q22.** (a) wins by a long way. Rating more often over the same 20 weeks
shrinks the standard error from `r f1(8 / sqrt(Sxx), 3)` to about
`r f1(8 / sqrt(sum((seq(0, 20, length.out = 60) - 10)^2)), 3)`, roughly the
familiar $\sqrt{n}$ rule: nearly three times the data, only about 1.6 times
the precision. Extending therapy to 60 weeks shrinks it to about
`r f1(8 / sqrt(sum(((0:59) - mean(0:59))^2)), 3)`, nearly five times more
precise, because a slope is best pinned down by points far apart in time.

The catch: this assumes the world stays linear for 60 weeks (Q5 says it
cannot, forever), and that ratings three times a week are independent of each
other (Section 3 says that is exactly what the linear world assumes, and
exactly what real mood data tend to violate).

**Q23.** Every fitted line passes through the average week (week 10) at that
rerun's average mood, so the lines pivot there, like a seesaw. A small error
in the slope barely moves the line near week 10 but is magnified further out.
We know Albert's expected mood mid-therapy far better than at intake, or next
week. Sections 10 and 11 both follow from this.

</details>

::: {.tool}
<strong class="tag">The tools you just used</strong>

**Simulation from a fitted model** (a parametric bootstrap, informally), the
**sampling distribution** $\hat\beta_1 \sim \mathcal{N}(\beta_1, \sigma^2 / S_{xx})$,
and the **standard error** as its standard deviation. **Study design** enters
through $S_{xx}$.
:::

::: {.question}
**Q24.** $\hat\sigma$ is itself an estimate. Use the reruns idea: would you
expect it to land exactly on the $\sigma$ of the world? What does that mean
for the standard error `lm()` reports?
:::

<details>
<summary>Show the answer</summary>

No. With 19 degrees of freedom, $\hat\sigma$ varies from rerun to rerun by
roughly ±1.3 around a true $\sigma$ of 8 (and is very slightly biased
downwards: $\hat\sigma^2$ is unbiased for $\sigma^2$, but its square root
averages a little under $\sigma$). So the standard error `lm()` reports,
computed from $\hat\sigma$, is itself uncertain. That is why intervals for the
slope use a $t$ distribution rather than a normal: the $t$ is the price of
not knowing $\sigma$.

</details>

---

# 9. How sure can we be? What "95%" means

From Albert's data, `lm()` gives a 95% confidence interval for his weekly
change of **`r f1(ci_b1[1], 2)` to `r f1(ci_b1[2], 2)`**. What does "95%"
promise?

It promises something about the **recipe**, not about this interval. To see
it, we need a world whose dials we *know*, so we can check each interval.
Below, the world's true weekly change is set to Albert's estimate,
`r f1(b1_hat, 2)` (the red line), and we rerun the study 2000 times.

```{r fig-coverage, echo=FALSE}
inputPanel(
  sliderInput("ci_sigma", "Size of weekly shocks (sigma):",
              min = 2, max = 20, value = round(s_hat), step = 1, width = "240px"),
  sliderInput("ci_level", "Confidence level:", min = 0.50, max = 0.99,
              value = 0.95, step = 0.01, width = "240px"),
  actionButton("ci_go", "Rerun 100 more studies")
)

ci_sims <- reactive({
  input$ci_go
  s  <- sim_studies(2000, input$ci_sigma)
  tq <- qt(1 - (1 - input$ci_level) / 2, n - 2)
  s$lo <- s$b1 - tq * s$se
  s$hi <- s$b1 + tq * s$se
  s
})

renderPlot({
  s <- ci_sims(); truth <- b1_hat
  cover <- s$lo <= truth & s$hi >= truth
  K <- 100
  base_par(mar = c(4.4, 4.6, 3.4, 1.2))
  xr <- truth + c(-1, 1) * max(1.2, 1.4 * max(abs(c(s$lo[1:K], s$hi[1:K]) - truth)))
  plot(NA, xlim = xr, ylim = c(0, K + 1), axes = FALSE,
       xlab = "interval for the weekly change", ylab = "study number")
  abline(v = pretty(xr, 6), col = COL$grid)
  abline(v = truth, col = COL$truth, lwd = 3)
  abline(v = 0, col = COL$muted, lty = 3, lwd = 1.5)
  for (i in 1:K) {
    hit <- cover[i]
    cc  <- if (hit) alpha(COL$model, 0.5) else COL$ink
    segments(s$lo[i], i, s$hi[i], i, col = cc, lwd = if (hit) 2 else 3.5)
    points(s$b1[i], i, pch = 19, cex = if (hit) 0.6 else 1.0, col = cc)
  }
  axis(1, col = COL$grid, col.ticks = COL$muted)
  axis(2, at = c(1, 25, 50, 75, 100), col = COL$grid, col.ticks = COL$muted)
  title(sprintf("%d%% intervals from 100 reruns", round(100 * input$ci_level)),
        adj = 0, line = 2.0, cex.main = 1.3, col.main = COL$ink)
  title(sprintf(
    "bold black = missed the true value (%d of these 100;  %.1f%% over 2000)",
    sum(!cover[1:K]), 100 * mean(cover)),
    adj = 0, line = 0.6, cex.main = 0.98, col.main = COL$muted, font.main = 1)
}, height = 460)
```

::: {.question}
**Q25.** Is it right to say *"there is a 95% probability that Albert's true
weekly change lies between `r f1(ci_b1[1], 2)` and `r f1(ci_b1[2], 2)`"*?

**Q26.** Drop the confidence level to 0.50 and rerun a few times. The intervals
get much shorter. Is a shorter interval a better answer?

**Q27.** Raise $\sigma$ to 20. Coverage stays near 95%. So what has gone
wrong?
:::

<details>
<summary>Show the answers</summary>

**Q25.** Not quite. Albert's true $\beta_1$ (if he has one) is a fixed
number; this interval either contains it or it does not. The 95% describes the
long-run behaviour of the recipe: across many studies, about 95 in 100
intervals catch the truth, which is what the figure shows. A defensible
sentence: *"Albert's mood improved by an estimated `r f1(b1_hat)` points per
week (95% CI `r f1(ci_b1[1])` to `r f1(ci_b1[2])`)."* And note the fine print:
this is 95% **if the linear world is the right world**. If the mechanism is
different, for example if weeks carry over into each other, the recipe's
promise no longer holds.

**Q26.** No. At 50% the intervals miss the truth half the time. You can have
any width you like by accepting being wrong more often. The only honest way to
a narrower interval is better data: smaller shocks, more ratings, a longer span.

**Q27.** The intervals are now very wide, typically about
±`r f1(qt(0.975, n - 2) * 20 / sqrt(Sxx))` points a week, and many reach down
to zero (the dotted line). The procedure is still *valid* (right 95% of the
time) but no longer *useful*: an interval from about zero to three points a
week is consistent with almost no change and with rapid improvement. For a
very changeable person, 21 weeks may simply not be enough to tell.

</details>

::: {.tool}
<strong class="tag">The tool you just used</strong>

The **confidence interval**
$\hat\beta_1 \pm t_{0.975,\,n-2} \cdot \mathrm{SE}(\hat\beta_1)$ and its
frequentist meaning: **coverage** is a property of the procedure, and it is
guaranteed only if the model is the right one.
:::

---

# 10. His expected mood versus next week's rating

Two questions that sound alike and are not:

- **Where is Albert's expected mood at week $t$?** This is uncertainty about the
  line: the **confidence band**.
- **Where will one actual week land?** This is uncertainty about the line
  **plus** next week's shock: the **prediction band**.

$$\mathrm{SE}(\text{expected mood at } t) = \sigma\sqrt{\tfrac{1}{n} + \tfrac{(t-\bar t)^2}{S_{xx}}}
\qquad
\mathrm{SE}(\text{one actual week at } t) = \sigma\sqrt{1 + \tfrac{1}{n} + \tfrac{(t-\bar t)^2}{S_{xx}}}$$

The extra **1** is variability (A). More data shrinks everything else. It
cannot shrink that.

```{r fig-bands, echo=FALSE}
inputPanel(
  sliderInput("bd2_n", "Ratings collected over the 20 weeks:",
              min = 5, max = 200, value = 21, step = 1, width = "300px"),
  checkboxInput("bd2_conf", "Confidence band (expected mood)", TRUE),
  checkboxInput("bd2_pred", "Prediction band (one actual week)", TRUE)
)

renderPlot({
  nn <- input$bd2_n
  if (nn == 21) {
    wk <- albert$week; y <- albert$mood
  } else {
    wk <- seq(0, 20, length.out = nn)
    y  <- with_seed(99, b0_hat + b1_hat * wk + rnorm(nn, 0, s_hat))
  }
  f2 <- lm(y ~ wk)
  nd <- data.frame(wk = seq(0, 20, length.out = 200))
  pc <- predict(f2, nd, interval = "confidence")
  pp <- predict(f2, nd, interval = "prediction")

  base_par()
  panel(c(-0.6, 20.6), c(10, 105), xat = seq(0, 20, 5),
        main = "Where is his expected mood?  vs.  Where will one week land?",
        sub  = sprintf("%d ratings%s   width at week 20:  confidence %.1f,  prediction %.1f",
                       nn, if (nn == 21) " (Albert)" else " (simulated)",
                       diff(pc[200, 2:3]), diff(pp[200, 2:3])))
  if (isTRUE(input$bd2_pred))
    polygon(c(nd$wk, rev(nd$wk)), c(pp[, "lwr"], rev(pp[, "upr"])),
            col = alpha(COL$model_lt, 0.40), border = NA)
  if (isTRUE(input$bd2_conf))
    polygon(c(nd$wk, rev(nd$wk)), c(pc[, "lwr"], rev(pc[, "upr"])),
            col = alpha(COL$model, 0.32), border = NA)
  lines(nd$wk, pc[, "fit"], col = COL$model, lwd = 3.5)
  points(wk, y, pch = 19, cex = if (nn > 60) 0.8 else 1.5,
         col = alpha(COL$ink, if (nn > 60) 0.45 else 1))
  legend("topleft", bty = "n", border = NA,
         fill = c(alpha(COL$model, 0.32), alpha(COL$model_lt, 0.40)),
         legend = c("95% confidence band", "95% prediction band"),
         text.col = COL$ink, cex = 1.05)
}, height = 420)
```

For **week 21**, the next session:

```{r week21, echo=TRUE}
predict(fit, data.frame(week = 21), interval = "confidence")  # his expected mood
predict(fit, data.frame(week = 21), interval = "prediction")  # one actual week
```

::: {.question}
**Q28.** Imagine Albert had rated his mood on his phone every day or two, 200
ratings over the same 20 weeks. Drag the slider up. What happens to each band?

**Q29.** Nora wants a rule: *"If next week's rating is below __, I will raise
it with Albert as a possible setback."* Which band, and what number? Which
band would a researcher comparing how fast clients improve under two
therapies need?
:::

<details>
<summary>Show the answers</summary>

**Q28.** The confidence band shrinks towards the line: with enough ratings,
the expected trajectory is known almost exactly. The prediction band barely
moves; it settles at about ±2σ and stays there. More data buys certainty
about **Albert's expected mood**, almost nothing about **how any one week
will go**.

**Q29.** The prediction band, because she is asking about one future rating
from one person: its lower limit for week 21 is about
**`r round(pi21[, "lwr"])`**. (Using the confidence band's lower limit,
about `r round(ci21[, "lwr"])`, she would "detect a setback" in many
perfectly ordinary weeks.) The researcher needs the confidence band, or the
interval for the slope, because they are asking about *average* change.
Quoting a narrow confidence interval as though it described what individual
clients can expect makes a treatment sound far more predictable than it is.

And the rule inherits every claim of the linear world: that Albert keeps
improving at the same rate, and that last week tells you nothing about next
week. The line reaches 100 around week `r round(hits_100)`, so the rule
certainly cannot be used for long.

</details>

::: {.tool}
<strong class="tag">The tools you just used</strong>

The **confidence interval for the mean response** versus the **prediction
interval for a new observation**, and the danger of **extrapolation**.
:::

---

# 11. Where you put zero

Back to Q16: a lower start and a steeper climb fitted almost as well. That
trade-off is a consequence of the pivot in Section 8. The intercept is the
line's height at week 0, the very edge of the data, ten weeks from the pivot,
so it catches the full swing of any error in the slope.

But "week 0" is our choice. **Slide the zero.** Nothing about Albert changes,
only where we read off his level.

```{r fig-centring, echo=FALSE}
inputPanel(
  sliderInput("cz_zero", "Read off Albert's level at week:",
              min = 0, max = 20, value = 0, step = 1, width = "320px")
)

renderPlot({
  z  <- input$cz_zero
  wz <- albert$week - z
  fz <- lm(albert$mood ~ wz)
  V  <- vcov(fz)
  rho <- V[1, 2] / sqrt(V[1, 1] * V[2, 2])
  ell <- conf_ellipse(V, coef(fz), df = n - 2)

  par(mfrow = c(1, 2)); base_par()

  panel(c(-0.6, 20.6), c(20, 95), xat = seq(0, 20, 5),
        main = "Same data, same line",
        sub  = sprintf("level at week %d:  %.1f +/- %.1f",
                       z, coef(fz)[1], 1.96 * sqrt(V[1, 1])))
  abline(fit, col = COL$model, lwd = 3.5)
  points(albert$week, albert$mood, pch = 19, cex = 1.4,
         col = alpha(COL$ink, 0.35))
  mu <- coef(fz)[1]; seI <- sqrt(V[1, 1])
  segments(z, mu - 1.96*seI, z, mu + 1.96*seI, col = COL$resid, lwd = 6)
  points(z, mu, pch = 19, cex = 2, col = COL$resid)
  abline(v = mean(albert$week), col = COL$muted, lty = 3, lwd = 2)
  text(mean(albert$week), 93, "pivot (mean week)", col = COL$muted,
       cex = 1.0, pos = if (z > 10) 2 else 4)
  text(z, 24, sprintf("SE of level = %.2f", seI), col = COL$resid,
       font = 2, cex = 1.1, pos = if (z > 10) 2 else 4)

  plot(NA, xlim = coef(fz)[1] + c(-10.5, 10.5),
       ylim = coef(fz)[2] + c(-1.1, 1.1),
       axes = FALSE, xlab = sprintf("level at week %d", z),
       ylab = "change per week")
  abline(h = pretty(coef(fz)[2] + c(-1.1, 1.1), 5),
         v = pretty(coef(fz)[1] + c(-10.5, 10.5), 5), col = COL$grid)
  polygon(ell, col = alpha(COL$model, 0.25), border = COL$model, lwd = 2.5)
  points(coef(fz)[1], coef(fz)[2], pch = 19, cex = 1.5, col = COL$resid)
  axis(1, col = COL$grid, col.ticks = COL$muted)
  axis(2, col = COL$grid, col.ticks = COL$muted)
  title("Level and change, jointly", adj = 0, line = 1.8,
        cex.main = 1.25, col.main = COL$ink)
  title(sprintf("correlation between the two estimates = %+.2f", rho),
        adj = 0, line = 0.5, cex.main = 0.98, col.main = COL$muted,
        font.main = 1)
}, height = 420)
```

::: {.question}
**Q30.** Slide from week 0 to week 20. Where is Albert's level pinned down
best, and where is the ellipse upright? Why there?

**Q31.** Does the uncertainty about the *weekly change* (the vertical extent of
the ellipse) change as you slide?
:::

<details>
<summary>Show the answers</summary>

**Q30.** At week 10, the mean week and the pivot. The standard error of the
level falls from `r f1(sqrt(vcov(fit)[1, 1]), 2)` at intake to
`r f1(sqrt(vcov(lm(mood ~ I(week - 10)))[1, 1]), 2)` at week 10, and the two
estimates become uncorrelated. Before week 10 the correlation is negative
(a higher level goes with a flatter climb, the trade-off from Q16); after it,
positive.

**Q31.** No. Moving zero changes where you read the line's height, not how
steep it is. **Uncertainty about the level depends on where you choose to
measure it; uncertainty about the rate of change is a real property of the
data.** In R, reading the level at mid-therapy is just **centring**:
`lm(mood ~ I(week - 10))`. In the multilevel and dynamic models later in this
course, where each person's own level matters, where you put zero decides
what that number means and how well it is estimated.

</details>

::: {.tool}
<strong class="tag">The tools you just used</strong>

**Centring** a predictor; the **covariance of the coefficient estimates**,
`vcov(fit)`; the **joint confidence region** for $(\beta_0, \beta_1)$.
:::

---

# 12. What Lecture 1 lets us say about Albert

::: {.question}
**Q32.** Before opening the box, write a short paragraph (at most 150 words)
for Nora: what does the linear model say about Albert, how sure can she be,
what should she expect next? Then add one sentence about what the model
*assumes* that you are not sure is true.
:::

<details>
<summary>Show one version</summary>

<div class="report">

*If Albert's mood follows a straight-line trend with independent weekly
shocks,* his expected mood rose by about **`r f1(b1_hat)` points per week**
(95% CI `r f1(ci_b1[1])` to `r f1(ci_b1[2])`), from about `r round(b0_hat)` at
intake to about `r round(fit_20)` at week 20. A trend this strong would be
very unlikely in a world where he was not changing ($p <$ .001). Individual
weeks scatter around that trend by about ±`r round(s_hat)` points; week 14,
`r round(-wk14_res)` points below trend, is the kind of week that turns up in
about one course of therapy in `r round(1 / p_any21)`. Next week's rating
would be expected around `r round(pi21[, "fit"])`, anywhere from roughly
`r round(pi21[, "lwr"])` to `r round(pi21[, "upr"])`.

*What I am not sure of:* that his weeks are independent. If a bad week
carries over into the next, these intervals are too narrow.

</div>

</details>

<div class="keybox">

**The model of Lecture 1**

$$\text{Mood}_t \;=\; \beta_0 + \beta_1\,t + \varepsilon_t,
\qquad \varepsilon_t \sim \mathcal{N}(0, \sigma)\ \text{independently}$$

| | As a mechanism | Albert's estimate |
|---|---|---|
| $\beta_0$ | his expected mood at intake | `r f1(b0_hat)` |
| $\beta_1$ | steady change per week, driven by time | `r f1(b1_hat, 2)` |
| $\sigma$ | size of life's memoryless weekly shocks: **variability (A)** | `r f1(s_hat, 2)` |
| $\mathrm{SE}(\hat\beta_1)$ | not part of the mechanism: how much one 21-week series tells us, **variability (B)** | `r f1(se_b1, 3)` |

**The linear world can produce data like Albert's** (Section 4). That is not
the same as being his mechanism. It says change is steady and never ends,
that it is driven by time rather than by Albert's own state, that
fluctuations leave no trace, and that his unpredictability never changes.
**Next lecture** we drop the third claim and let this week depend on last
week.

</div>

---

# 13. Exercises

**E1. Does a bad week spill into the next?**
Run `acf(resid(fit))` and look at lag 1. If Albert's residuals were
*positively* correlated from one week to the next, would the standard error
`lm()` reports be too large or too small? Why is that the dangerous direction?

<details>
<summary>Hint</summary>

Here the lag-1 autocorrelation is about
`r f1(acf(resid(fit), plot = FALSE)$acf[2], 2)`: small, and with 21 weeks
hard to distinguish from zero. With positive carry-over, consecutive weeks
partly repeat each other, so 21 weeks hold *less* independent information than
21 separate observations would. `lm()` counts all 21 fully, believes it knows
more than it does, and reports a standard error that is **too small**:
intervals too narrow, p-values too low, overconfident conclusions. Carry-over
is the rule, not the exception, in intensive repeated measures.

</details>

**E2. Simulate the violation.**
Adapt the one-line recipe from Section 1 so that each week's shock carries
over half of last week's shock:
`eps[t] <- 0.5 * eps[t - 1] + rnorm(1, 0, sigma)`. Simulate 2000 Alberts
from this world, fit `lm(mood ~ week)` to each, and compare (a) the actual
spread of the slopes with (b) the average standard error `lm()` reports. What
do you find, and what does it mean for Q25?

**E3. A more changeable Albert.**
In Section 8, predict before you drag: if $\sigma$ goes from 8 to 12, by what
factor does the spread of estimated slopes grow? How long would you need to
observe a very changeable client before you could say whether therapy is
working?

**E4. From time to state.**
The linear model says *time* drives Albert's mood. A dynamic model says *last
week's mood* does. Fit

```{r e4, eval=FALSE, echo=TRUE}
m  <- albert$mood
wk <- albert$week
summary(lm(m[-1] ~ m[-21]))              # this week from last week
summary(lm(m[-1] ~ m[-21] + wk[-1]))     # ... and from the trend as well
```

What does the coefficient on last week's mood mean in the first model? What
happens to it in the second, and why?

<details>
<summary>Hint</summary>

In the first model it is about 0.66, which looks like strong *inertia*, mood
carrying over from week to week. With the trend included it collapses to
about −0.2 and is far from significant. Because Albert improves steadily, last
week's mood tells you roughly how far into therapy he is, and so roughly how
good next week will be: the "inertia" was the trend in disguise. Telling a
trend apart from genuine carry-over is one of the central problems of
dynamic modelling, and the starting point of next lecture.

</details>

---

<p style="color:#898781; font-size:0.9em; margin-top:3em;">
Little Albert refers to Watson, J. B., &amp; Rayner, R. (1920). Conditioned
emotional reactions. <em>Journal of Experimental Psychology, 3</em>(1), 1–14.
The grown-up Albert, his therapist and his mood ratings are fictional.
Rendered with R and Shiny; no packages beyond <code>shiny</code> are required.
</p>
