---
title: "Lecture 2 · Albert among Others"
subtitle: "Individual differences, pooling, and the multilevel model"
author: "Dynamic Systems Thinking for Psychologists"
date: "`r format(Sys.Date(), '%d %B %Y')`"
output:
  html_document:
    theme: flatly
    toc: true
    toc_float:
      collapsed: false
    toc_depth: 1
    df_print: kable
runtime: shiny
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE,
                      fig.align = "center")
library(shiny)
library(lme4)
```

```{css, echo=FALSE}
body { font-size: 16px; }
h1 { margin-top: 1.6em; }
details {
  background: #f4f7fb;
  border-left: 4px solid #2a78d6;
  padding: 0.6em 1em;
  margin: 1em 0 1.6em 0;
  border-radius: 4px;
}
details > summary { cursor: pointer; font-weight: 600; color: #1c5cab; }
details[open] > summary { margin-bottom: 0.8em; }
.question {
  background: #fff8f2;
  border-left: 4px solid #eb6834;
  padding: 0.8em 1em;
  margin: 1.4em 0 0.4em 0;
  border-radius: 4px;
}
.question p:first-child { margin-top: 0; }
.question p:last-child  { margin-bottom: 0; }
.keybox {
  background: #eef4fc;
  border: 1px solid #b7d3f6;
  padding: 1em 1.2em;
  border-radius: 6px;
  margin: 1.4em 0;
}
.tool {
  border-top: 1px solid #d6d4cc;
  border-bottom: 1px solid #d6d4cc;
  padding: 0.6em 0;
  margin: 1.4em 0;
  color: #3d3c38;
}
.tool > p:first-child { margin-top: 0; }
.tool strong.tag {
  font-variant: small-caps; letter-spacing: 0.04em; color: #898781;
}
.report {
  background: #fbfaf6;
  border: 1px solid #e1e0d9;
  border-left: 4px solid #0b0b0b;
  font-family: Georgia, serif;
  padding: 1em 1.3em;
  margin: 1.4em 0;
}
.part {
  margin: 2.4em 0 0.4em 0;
  padding: 0.5em 0.9em;
  background: #0b0b0b;
  color: #fcfcfb;
  font-weight: 700;
  letter-spacing: 0.03em;
  border-radius: 4px;
}
.shiny-input-container { margin-bottom: 4px; }
```



```{r data, echo=TRUE}
# Albert's 21 weekly mood ratings, exactly as in Lecture 1.
albert <- data.frame(week = 0:20,
                     mood = c(46, 35, 46, 46, 43, 30, 46, 45, 56, 53, 55,
                              54, 59, 61, 44, 77, 73, 71, 64, 78, 71))
```

```{r clinic-data, echo=FALSE}
# The other 29 clients of Albert's clinic. They are simulated from a two-level
# story (Section 1) with fixed dials and a fixed seed, so everyone gets the
# same clinic. Keep the dials to yourself until Section 8.
make_clinic <- function(seed) {
  set.seed(seed)
  J <- 29
  g00 <- 40; g10 <- 1.0; t0 <- 7; t1 <- 0.5; rho <- -0.3; sig <- 8
  S <- matrix(c(t0^2, rho * t0 * t1, rho * t0 * t1, t1^2), 2)
  L <- t(chol(S)); u <- t(L %*% matrix(rnorm(2 * J), 2))
  b0 <- g00 + u[, 1]; b1 <- g10 + u[, 2]
  nobs <- sample(c(rep(21, 20), sample(4:10, 9, replace = TRUE)))
  d <- do.call(rbind, lapply(1:J, function(j) {
    wk <- 0:(nobs[j] - 1)
    data.frame(client = j, week = wk,
               mood = pmin(pmax(round(b0[j] + b1[j] * wk +
                                      rnorm(length(wk), 0, sig)), 0), 100))
  }))
  list(d = d, b0 = b0, b1 = b1)
}
cl_sim <- make_clinic(30)
others <- cl_sim$d; others$client <- others$client + 1
clinic <- rbind(data.frame(client = 1, albert), others)
TRUE_B1_OTHERS <- cl_sim$b1          # true slopes of clients 2-30 (not Albert)
TRUE_B0_OTHERS <- cl_sim$b0

ids   <- sort(unique(clinic$client))
label <- paste("Client", ids); label[1] <- "Albert"; label[6] <- "Jonas"; label[14] <- "Lena"
clinic$name <- factor(label[clinic$client], levels = label)
n_i   <- as.vector(table(clinic$client))
ID_A <- 1; ID_J <- 6; ID_L <- 14
```

```{r helpers, echo=FALSE}
COL <- list(
  ink = "#0b0b0b", muted = "#898781", grid = "#e1e0d9",
  model = "#2a78d6", model_lt = "#9ec5f4", resid = "#eb6834",
  truth = "#e34948", dark = "#104281", up = "#2f9e6e", hi = "#8a4fc9"
)
alpha <- function(col, a) rgb(t(col2rgb(col)), alpha = a * 255, maxColorValue = 255)

base_par <- function(...)
  par(mar = c(4.4, 4.6, 3.2, 1.2), mgp = c(2.8, 0.7, 0), las = 1,
      col.axis = COL$muted, fg = COL$muted, tcl = -0.3,
      cex.lab = 1.1, cex.axis = 1.0, ...)

panel <- function(xlim, ylim, xlab = "Week of therapy", ylab = "Mood (0-100)",
                  main = "", sub = "", xat = NULL, yat = NULL) {
  plot(NA, xlim = xlim, ylim = ylim, xlab = xlab, ylab = ylab, axes = FALSE)
  if (is.null(xat)) xat <- pretty(xlim, 6)
  if (is.null(yat)) yat <- pretty(ylim, 6)
  abline(h = yat, v = xat, col = COL$grid, lwd = 1)
  axis(1, at = xat, col = COL$grid, col.ticks = COL$muted)
  axis(2, at = yat, col = COL$grid, col.ticks = COL$muted)
  if (nzchar(main))
    title(main = main, adj = 0, line = 1.8, cex.main = 1.3,
          col.main = COL$ink, font.main = 2)
  if (nzchar(sub))
    title(main = sub, adj = 0, line = 0.5, cex.main = 0.98,
          col.main = COL$muted, font.main = 1)
}

# The two-level story: J clients, each with their own dials, all observed on
# the same weeks. Returns the dials and a (weeks x J) matrix of ratings.
two_level_world <- function(J, g00, g10, tau0, tau1, rho, sigma, weeks = 0:20) {
  z0 <- rnorm(J); z1 <- rho * z0 + sqrt(1 - rho^2) * rnorm(J)
  b0 <- g00 + tau0 * z0; b1 <- g10 + tau1 * z1
  Y  <- outer(weeks, b1) + matrix(b0, length(weeks), J, byrow = TRUE) +
        matrix(rnorm(length(weeks) * J, 0, sigma), length(weeks), J)
  list(b0 = b0, b1 = b1, Y = Y, weeks = weeks)
}

# Each client's own least-squares slope, for clients observed on the same weeks.
ols_slopes <- function(Y, weeks) {
  xc <- weeks - mean(weeks)
  as.vector(crossprod(xc, Y)) / sum(xc^2)
}

Sxx_n <- function(n) n * (n^2 - 1) / 12      # spread of weeks 0, 1, ..., n-1

dials2 <- function(id, g10 = 1, tau0 = 7, tau1 = 0.5, rho = -0.3, sigma = 8)
  list(
    sliderInput(paste0(id, "_g10"), "Typical change per week (gamma10):",
                min = -1, max = 3, value = g10, step = 0.1, width = "230px"),
    sliderInput(paste0(id, "_t0"), "Spread of starting levels (tau0):",
                min = 0, max = 15, value = tau0, step = 0.5, width = "230px"),
    sliderInput(paste0(id, "_t1"), "Spread of change per week (tau1):",
                min = 0, max = 2, value = tau1, step = 0.05, width = "230px"),
    sliderInput(paste0(id, "_rho"), "Start vs. change correlation (rho):",
                min = -0.9, max = 0.9, value = rho, step = 0.1, width = "230px"),
    sliderInput(paste0(id, "_sg"), "Size of weekly shocks (sigma):",
                min = 0, max = 20, value = sigma, step = 0.5, width = "230px")
  )
```

```{r numbers, echo=FALSE}
f1 <- function(x, d = 1) formatC(x, format = "f", digits = d)
pct <- function(x) round(100 * x)

fit_A  <- lm(mood ~ week, data = albert)                 # Lecture 1
b1_A   <- unname(coef(fit_A)[2]); b0_A <- unname(coef(fit_A)[1])
ci_A   <- confint(fit_A)["week", ]

# complete pooling
m_cp   <- lm(mood ~ week, data = clinic)
cp     <- coef(m_cp); se_cp <- summary(m_cp)$coefficients["week", "Std. Error"]

# no pooling: each client's own line
np     <- t(sapply(split(clinic, clinic$client),
                   function(x) coef(lm(mood ~ week, data = x))))
np_se  <- sapply(split(clinic, clinic$client), function(x)
                 summary(lm(mood ~ week, data = x))$coefficients["week", "Std. Error"])

# partial pooling: the multilevel model
m_re   <- lmer(mood ~ week + (week | client), data = clinic)
g_hat  <- fixef(m_re); se_g <- sqrt(diag(vcov(m_re)))
vc     <- VarCorr(m_re)$client
tau0_h <- attr(vc, "stddev")[1]; tau1_h <- attr(vc, "stddev")[2]
rho_h  <- attr(vc, "correlation")[1, 2]; sig_h <- sigma(m_re)
pp     <- as.matrix(coef(m_re)$client)                   # each client's pooled line

pct_A_pp <- pnorm((pp[ID_A, 2] - g_hat[2]) / tau1_h)
pct_A_np <- pnorm((np[ID_A, 2] - g_hat[2]) / tau1_h)
w_n      <- function(n, t1 = tau1_h, s = sig_h) t1^2 / (t1^2 + s^2 / Sxx_n(n))
n_half   <- min(which(sapply(2:60, w_n) > 0.5)) + 1

rmse_np <- sqrt(mean((np[-1, 2] - TRUE_B1_OTHERS)^2))
rmse_pp <- sqrt(mean((pp[-1, 2] - TRUE_B1_OTHERS)^2))
```

# How to use this document

Open it in RStudio and click **Run Document** (or run
`rmarkdown::run("lecture02_individual_differences.Rmd")`). It needs the
`shiny` and `lme4` packages.

- **Orange boxes** are questions. Answer them, in writing, *before* opening the
  blue box underneath.
- **Sliders** are there to be moved.
- **Grey-ruled boxes** name the statistical tool you have just used.

---

# Where we are: is therapy going well for Albert?

In Lecture 1 we told the simplest story about Albert's mood: therapy moves his
expected mood along a straight line, and every week life adds a shock. Fitted
to his 21 ratings, the story says his mood rises by about
**`r f1(b1_A, 2)` points per week** (95% CI `r f1(ci_A[1], 2)` to
`r f1(ci_A[2], 2)`).

```{r fig-albert, echo=FALSE}
renderPlot({
  base_par()
  panel(c(-0.6, 20.6), c(20, 90), xat = seq(0, 20, 5),
        main = "Albert's 21 weeks and his line from Lecture 1",
        sub = sprintf("mood = %.1f + %.2f x week", b0_A, b1_A))
  abline(fit_A, col = COL$model, lwd = 3.5)
  points(albert$week, albert$mood, pch = 19, cex = 1.7, col = COL$ink)
}, height = 360)
```

His therapist Nora now wants to know something Lecture 1 cannot answer:

> **Is therapy going *well* for Albert?**

::: {.question}
**Q1.** Albert's mood rises by about `r f1(b1_A, 1)` points a week. Is that
good? Write down what you would need to know to answer.
:::

<details>
<summary>Show the answer</summary>

On its own, a slope is just a number. To call it good, slow or worrying you
need a **reference**: how fast do clients like Albert, in this therapy, with
this questionnaire, usually improve? And how much do they differ? A weekly
gain of `r f1(b1_A, 1)` is exceptional if nearly everyone gains about 0.5, but
ordinary if clients range from −1 to +4.

That is exactly what clinics do in *routine outcome monitoring*: a client's
progress is compared with the expected recovery curves of similar clients, and
those who fall off track are flagged. To do it, we need a story that is not
just about Albert but about **clients in general**, and about how Albert
relates to them.

</details>

This lecture tells that story. Part 1 builds and explores a **two-level
story**: a story about how each client's dials come about, as well as how
each week's rating comes about. Part 2 fits it to Albert's clinic, step by
step: one line for everyone, a separate line for each client, and finally the
multilevel model, which does both at once.

---

<div class="part">Part 1 · A clinic of Alberts</div>

# 1. A story with two levels

Lecture 1's story had three dials, $\beta_0$, $\beta_1$ and $\sigma$, and
they belonged to Albert. But every client has their own. Some start lower,
some improve faster, some hardly change at all. So the story needs a second
level.

**Level 1: how a client's weeks go.** Exactly Lecture 1's story, now for
client $i$, with that client's own dials:

$$\text{Mood}_{it} = \beta_{0i} + \beta_{1i}\, t + \varepsilon_{it},
\qquad \varepsilon_{it} \sim \mathcal{N}(0, \sigma)$$

**Level 2: where a client's dials come from.** Each client's starting level
and rate of change are drawn from the clinic's *population* of clients:

$$\beta_{0i} \sim \mathcal{N}(\gamma_{00}, \tau_0), \qquad
\beta_{1i} \sim \mathcal{N}(\gamma_{10}, \tau_1)$$

with a correlation $\rho$ between the two (do clients who start lower tend to
improve faster?).

As code, it is Lecture 1's loop with an outer loop wrapped around it:

```{r recipe2, eval=FALSE, echo=TRUE}
for (i in 1:n_clients) {                       # level 2: who you are
  b0_i <- rnorm(1, g00, tau0)                  #   this client's starting level
  b1_i <- rnorm(1, g10, tau1)                  #   this client's rate of change
  for (t in 0:20) {                            # level 1: how your weeks go
    mood[i, t + 1] <- b0_i + b1_i * t + rnorm(1, 0, sigma)
  }
}
```

(For simplicity this sketch leaves out $\rho$; the widget below includes it.)

| Dial | What it claims |
|---|---|
| $\gamma_{00}$ | the typical client's expected mood at intake |
| $\gamma_{10}$ | the typical client's change per week |
| $\tau_0$ | how much clients differ in where they start |
| $\tau_1$ | how much clients differ in how fast they change |
| $\rho$ | whether starting level and rate of change go together |
| $\sigma$ | the size of life's weekly shocks, within a client |

**Open a clinic.** Each line is one client's expected mood; the dots are the
ratings life would produce. The right panel shows the population of rates of
change that clients are drawn from, and where a weekly gain like Albert's
would sit in it.

```{r fig-clinic-sim, echo=FALSE}
do.call(inputPanel, c(dials2("cs"), list(
  checkboxInput("cs_pts", "Show weekly ratings", FALSE),
  actionButton("cs_go", "Open a new clinic")
)))

cs_sim <- reactive({
  input$cs_go
  two_level_world(12, 40, input$cs_g10, input$cs_t0, input$cs_t1,
                  input$cs_rho, input$cs_sg)
})

renderPlot({
  w <- cs_sim(); wk <- w$weeks
  cols <- hcl.colors(12, "Dark 3")
  par(mfrow = c(1, 2)); base_par()

  yl <- range(c(0, 100, outer(wk, w$b1) + matrix(w$b0, 21, 12, byrow = TRUE)))
  panel(c(-0.6, 20.6), yl, xat = seq(0, 20, 5),
        main = "12 clients, own dials",
        sub = "lines = each client's expected mood")
  for (j in 1:12) {
    if (isTRUE(input$cs_pts))
      points(wk, w$Y[, j], pch = 19, cex = 0.7, col = alpha(cols[j], 0.6))
    abline(a = w$b0[j], b = w$b1[j], col = cols[j], lwd = 2.5)
  }

  g10 <- input$cs_g10; t1 <- max(input$cs_t1, 0.01)
  xr <- range(c(-1, 4, g10 + c(-3.5, 3.5) * t1))
  plot(NA, xlim = xr, ylim = c(0, dnorm(0, 0, t1) * 1.25), axes = FALSE,
       xlab = "a client's change per week (beta1)", ylab = "")
  abline(v = pretty(xr, 6), col = COL$grid)
  xs <- seq(xr[1], xr[2], length.out = 400)
  polygon(c(xs, rev(xs)), c(dnorm(xs, g10, t1), rep(0, 400)),
          col = alpha(COL$model_lt, 0.6), border = COL$model, lwd = 2)
  hi <- xs >= b1_A
  polygon(c(xs[hi], rev(xs[hi])), c(dnorm(xs[hi], g10, t1), rep(0, sum(hi))),
          col = alpha(COL$model, 0.7), border = NA)
  points(w$b1, rep(dnorm(0, 0, t1) * 0.04, 12), pch = 19, col = cols, cex = 1.4)
  abline(v = b1_A, col = COL$resid, lwd = 3, lty = 2)
  axis(1, col = COL$grid, col.ticks = COL$muted)
  title("Population of change rates", adj = 0, line = 1.8,
        cex.main = 1.3, col.main = COL$ink)
  title(sprintf("%.2f per week beats %.0f%% of clients",
                b1_A, 100 * pnorm(b1_A, g10, t1)),
        adj = 0, line = 0.5, cex.main = 0.98, col.main = COL$muted, font.main = 1)
}, height = 420)
```

::: {.question}
**Q2.** Set both spreads, $\tau_0$ and $\tau_1$, to 0. What does the clinic
look like? What story is this?

**Q3.** Now keep $\tau_1 = 0$ but set $\tau_0 = 10$. Clients differ, but in
what way? Is therapy "working equally well" for all of them?

**Q4.** With $\gamma_{10} = 1$ and $\tau_1 = 0.5$, how exceptional is a
weekly gain of `r f1(b1_A, 2)`? Now set $\tau_1 = 1.5$. What changed, and what
does that say about whether Albert's slope is "good"?

**Q5.** Set $\rho$ to −0.7. What pattern appears in the left panel? Why might
it be plausible for real clients?
:::

<details>
<summary>Show the answers</summary>

**Q2.** Every client has the same expected line; their charts differ only
through life's shocks. It is Lecture 1's story applied to everyone at once:
a clinic of identical Alberts. Any differences you see between clients'
charts are luck.

**Q3.** Parallel lines. Clients start at different levels but improve at the
same rate. Therapy adds the same amount per week for everyone, yet clients
who started low are still low after 20 weeks. Whether that counts as "working
equally well" depends on whether you care about change or about where people
end up, and that is a clinical question, not a statistical one.

**Q4.** With $\tau_1 = 0.5$, a gain of `r f1(b1_A, 2)` is faster than about
`r pct(pnorm(b1_A, 1, 0.5))`% of clients: exceptional. With $\tau_1 = 1.5$ it
beats only about `r pct(pnorm(b1_A, 1, 1.5))`%: good, not remarkable. The
same number for Albert means different things depending on **how much clients
differ**. "Good" is always relative to a population, and $\tau_1$ describes
that population as much as $\gamma_{10}$ does.

**Q5.** A fan that narrows over time: clients who start low climb faster and
catch up. Plausible reasons: people who start lower have more room to
improve; very low intake ratings are partly bad-luck weeks, which fade (the
regression-to-the-mean effect from Lecture 1); and the scale has a ceiling.

</details>

::: {.tool}
<strong class="tag">The tool you just used</strong>

A **two-level (multilevel, hierarchical) generative model**. Level 1:
$Y_{it} = \beta_{0i} + \beta_{1i} t + \varepsilon_{it}$. Level 2:
$(\beta_{0i}, \beta_{1i})$ drawn from a bivariate normal with means
$(\gamma_{00}, \gamma_{10})$, standard deviations $(\tau_0, \tau_1)$ and
correlation $\rho$. The $\tau$'s describe **between-person** variability;
$\sigma$ describes **within-person** variability.
:::

---

# 2. Where do the dials come from?

Section 1 drew every client's dials from a **normal**, bell-shaped population.
That was a choice, and in a generative story every choice is a claim about how
the world works. This one is a claim about **how clients come to differ**.

The bell is the right shape when a client's rate of change is the **sum of many
small, independent influences, none of which dominates**: whether they have a
supportive partner, a stable job, enough sleep, how well they click with their
therapist, and so on. Each influence nudges the rate a little up or down, and
the nudges add up. The **central limit theorem** says that such a sum ends up
close to a bell, whatever each nudge looks like on its own. It is the same
argument that makes body height roughly normal.

Each part of that sentence (*many*, *small*, *independent*, *adding up*) is a
substantive claim, and each can fail. The widget below makes 5000 clients'
dials from four different stories and compares them with the bell that has the
same mean and SD, which is the bell a multilevel model would assume.

- **Many small nudges that add up.** Each influence is either present (for 30%
  of clients) or absent, and a present influence adds a little to the
  client's rate of change.
- **Many small nudges that multiply.** A present influence scales the client's
  responsiveness up by a percentage, so influences amplify each other: a
  boost helps more in someone who is already improving.
- **Two kinds of clients.** Responders improve by about 1.5 points a week;
  non-responders don't change. One big cause instead of many small ones.
- **A bounded dial.** Some dials can only take values in a range. *Stickiness*,
  the share of last week's mood that carries over into this week, lies between
  0 (none) and 1 (all of it). Later lectures model this dial. Here the nudges
  add up on the log-odds scale, and the result is squeezed into (0, 1).

```{r mech-numbers, echo=FALSE}
# Four stories of how clients come to differ. Each of K influences is present
# for a client with probability 0.3 and, if present, has a size between 0.5
# and 1.5; z is the standardised sum of a client's influences.
mech_dials <- function(mech, K, p_resp, N = 5000) {
  nudges <- matrix(rbinom(N * K, 1, 0.3) * runif(N * K, 0.5, 1.5), N, K)
  z <- (rowSums(nudges) - 0.3 * K) / sqrt(0.235 * K)
  switch(mech,
         add   = 1 + 0.5 * z,
         mult  = exp(-0.47^2 / 2 + 0.47 * z),
         mix   = ifelse(runif(N) < p_resp, rnorm(N, 1.5, 0.3), rnorm(N, -0.1, 0.3)),
         bound = plogis(qlogis(0.8) + 1.2 * z))
}
bell_vs <- function(x, lo, hi = Inf) {
  m <- mean(x); s <- sd(x)
  c(obs = mean(x < lo | x > hi), bell = pnorm(lo, m, s) + pnorm(hi, m, s, lower.tail = FALSE))
}
set.seed(2)
mx_mult  <- mech_dials("mult", 50, 0.5, 2e5)
mx_mix85 <- mech_dials("mix", 50, 0.85, 2e5)
mx_bound <- mech_dials("bound", 50, 0.5, 2e5)
nb_mult_worse <- bell_vs(mx_mult, 0)
nb_mult_fast  <- bell_vs(-mx_mult, -2)                   # share above 2 per week
nb_mix_worse  <- bell_vs(mx_mix85, 0)
nb_bound      <- bell_vs(mx_bound, 0, 1)
```

```{r fig-mechanisms, echo=FALSE}
inputPanel(
  radioButtons("mx_mech", "How do clients come to differ?",
               choices = c("Many small nudges that add up" = "add",
                           "Many small nudges that multiply" = "mult",
                           "Two kinds of clients" = "mix",
                           "A bounded dial: stickiness" = "bound"),
               selected = "add", width = "280px"),
  sliderInput("mx_k", "Number of influences per client:",
              min = 1, max = 50, value = 3, step = 1, width = "230px"),
  sliderInput("mx_p", "Share of responders (two kinds only):",
              min = 0.05, max = 0.95, value = 0.5, step = 0.05, width = "230px"),
  actionButton("mx_go", "New clients")
)

renderPlot({
  input$mx_go
  mech <- input$mx_mech
  x <- mech_dials(mech, input$mx_k, input$mx_p)
  m <- mean(x); s <- sd(x)
  bound <- mech == "bound"
  xlab <- if (bound) "stickiness (share of last week's mood carried over)"
          else "a client's change per week"
  main <- switch(mech, add = "Nudges that add up", mult = "Nudges that multiply",
                 mix = "Two kinds of clients", bound = "A bounded dial")
  sub <- if (bound) {
    b <- bell_vs(x, 0, 1)
    sprintf("the bell puts %.1f%% of clients below 0 or above 1",
            100 * b["bell"])
  } else {
    w <- bell_vs(x, 0); f <- bell_vs(-x, -2)
    sprintf("worse: %.1f%% (bell %.1f%%)   faster than 2/week: %.1f%% (bell %.1f%%)",
            100 * w["obs"], 100 * w["bell"], 100 * f["obs"], 100 * f["bell"])
  }

  par(mfrow = c(1, 2)); base_par()
  xl <- if (bound) c(-0.2, 1.2) else range(c(-1, 3, m + c(-3.5, 3.5) * s))
  br <- seq(min(x, xl[1]) - 0.01, max(x, xl[2]) + 0.01, length.out = 70)
  h  <- hist(x, breaks = br, plot = FALSE)
  xs <- seq(xl[1], xl[2], length.out = 400)
  plot(NA, xlim = xl, ylim = c(0, max(h$density, dnorm(m, m, s)) * 1.1), axes = FALSE,
       xlab = xlab, ylab = "")
  abline(v = pretty(xl, 7), col = COL$grid)
  if (bound) rect(c(-1, 1), 0, c(0, 2), 1e3, col = alpha(COL$truth, 0.08), border = NA)
  plot(h, freq = FALSE, add = TRUE, col = alpha(COL$resid, 0.55), border = NA)
  lines(xs, dnorm(xs, m, s), col = COL$model, lwd = 3)
  if (!bound) abline(v = 0, col = COL$muted, lwd = 1.5, lty = 2)
  axis(1, col = COL$grid, col.ticks = COL$muted)
  legend(if (bound) "topleft" else "topright", bty = "n", pch = c(15, NA),
         pt.cex = 2, lwd = c(NA, 3), col = c(alpha(COL$resid, 0.55), COL$model),
         legend = c("5000 clients", "bell, same mean and SD"), cex = 0.95)
  title(main, adj = 0, line = 1.8, cex.main = 1.3, col.main = COL$ink)
  mtext(sub, side = 3, line = 0.4, adj = 0, cex = 0.9, col = COL$muted)

  q <- qqnorm(sample(x, 500), plot.it = FALSE)
  panel(range(q$x), range(q$y), xlab = "where a bell would put this client (SDs)",
        ylab = "client's dial", main = "Against the bell",
        sub = "on the line = bell-shaped")
  abline(m, s, col = COL$model, lwd = 3)
  points(q$x, q$y, pch = 19, cex = 0.8, col = alpha(COL$resid, 0.6))
}, height = 420)
```

::: {.question}
**Q6.** Start with *nudges that add up* and move the number of influences from
1 to 3, 10 and 50. When does the histogram start to look like the bell? Why
does it get there, even though a single influence (absent for most clients)
looks nothing like a bell?

**Q7.** Switch to *nudges that multiply* (50 influences). How does the shape
differ from the bell? Compare the share of clients getting worse, and of very
fast improvers, with what the bell says. What would therapy have to be like for
this story to be right?

**Q8.** Switch to *two kinds of clients*, with half responders, then with 85%.
At 85%, would you still spot the two kinds if the clinic had 30 clients rather
than 5000? What does the bell say about how many clients get worse, and why
does that matter for Nora?

**Q9.** Switch to *a bounded dial*. What does the bell claim about some
clients? How could you keep a bell in the story and still make it coherent?
:::

<details>
<summary>Show the answers</summary>

**Q6.** With one influence, most clients (the 70% without it) share exactly
the same rate and the rest form a separate block: two types of client, nothing
like a bell. With three it is lumpy and skewed to the right: about a third of clients
have none of the influences and pile up at the lowest rate, while only a few
have all three. By about 10 to 20
influences the histogram is hard to tell from the bell, and the dots in the
right panel lie on the line. That is the central limit theorem: when many
independent nudges are added, the shape of each single nudge washes out. The
bell is a good bet whenever you believe the many-small-additive-causes story,
and "many" need not be very many.

**Q7.** Skewed to the right: nobody gets worse, most clients sit a little
below the mean, and a long tail of clients improve very fast. The bell says
about `r f1(100 * nb_mult_worse["bell"])`% of clients get worse; in this story
`r f1(100 * nb_mult_worse["obs"])`% do. The bell says
`r f1(100 * nb_mult_fast["bell"])`% improve by more than 2 points a week; in
this story `r f1(100 * nb_mult_fast["obs"])`% do. The story fits a therapy in
which ingredients reinforce each other (hope makes engagement pay off more,
engagement makes new skills pay off more), and in which something can only
scale improvement up or down, never turn it into deterioration. The result is
a *lognormal* population: a bell on the log scale.

**Q8.** With half responders there are two humps, and the bell puts most
clients in the valley between them, where almost nobody is. With 85% there is
one hump with a long left tail. With 30 clients, and each client's slope
estimated with noise (Section 3), that would be very hard to tell from a
slightly lopsided bell. The bell says about `r f1(100 * nb_mix_worse["bell"])`%
of clients get worse; in this story it is `r f1(100 * nb_mix_worse["obs"])`%.
A clinic that sets its expectations from the bell would think deterioration is
rarer than it is. It also asks the wrong question: in this story the clinical
question is *who is a responder?*, and a single bell has no way to ask it. If
you know what distinguishes the groups (a diagnosis, a symptom profile), it
belongs in the model as a predictor; if you don't, a *mixture* model (a growth
mixture model) tries to find the groups.

**Q9.** The bell puts about `r f1(100 * nb_bound["bell"])`% of clients at
stickiness above 1 or below 0. Above 1, a mood would carry over more than
fully and grow week after week without limit: not a person, an explosion. The
fix is to keep the bell but on a **transformed scale**: put the bell on the
log-odds of stickiness and squeeze it back into (0, 1), which is exactly what
the widget did. Choosing that scale is itself a claim: nudges add on the
log-odds scale, so the same nudge matters less for a client whose stickiness
is already close to 0 or 1.

</details>

<div class="keybox">

**What the bell claims about people.**
Deviations above and below the typical client are equally common
(*symmetry*). Extreme clients are very rare: fewer than 3 in 1000 lie beyond
3 SDs (*thin tails*). Any value is possible (*no bounds*). And with two dials,
the relation between them is a straight line with the same scatter
everywhere, so a single correlation $\rho$ can describe it.

Whenever this lecture later says that Albert is faster than some percentage
of clients, that percentage is read off the bell.

</div>

**A second, weaker reading.** Sometimes there is no mechanism story. Then the
bell can be read as a statement about what we *know*, not about what the
clinic's population looks like: before seeing the data, we have no reason to
expect any particular client to be faster or slower than any other (the
clients are *exchangeable*), and we are only willing to commit to a centre and
a spread. Among all distributions with a given mean and SD, the normal is the
one that adds the least extra structure. That reading is defensible, but it
ends as soon as you know something that distinguishes clients, such as their
diagnosis or that they dropped out. That knowledge belongs in the story.

::: {.tool}
<strong class="tag">The tools you just used</strong>

The **central limit theorem**: many small, independent, additive causes give a
**normal** population. Multiplicative causes give a **lognormal** one (normal
on the log scale). A few large causes or discrete types give a **mixture**.
Bounded dials need a bell on a **transformed scale** (log-odds, log). The
**normal QQ plot** checks the shape of a sample against a bell. Behind any
population distribution sits a weaker assumption, **exchangeability**: before
seeing the data, nothing distinguishes one client from another.
:::

---

# 3. The trouble with comparing clients' slopes

Albert's `r f1(b1_A, 2)` is not his true rate of change. It is an
*estimate*, and Lecture 1 showed it would wobble if his weeks were rerun.
Every other client's slope wobbles too, and clients with only a few ratings
wobble a lot. So if Nora lines up the clinic's *estimated* slopes to see where
Albert stands, how much of the spread is real?

Below, 1000 clients are drawn from the two-level story. Blue: their **true**
rates of change. Orange: the rates Nora would **estimate** from each client's
own ratings.

```{r fig-true-vs-est, echo=FALSE}
inputPanel(
  sliderInput("te_t1", "True spread of change per week (tau1):",
              min = 0, max = 1.5, value = 0.5, step = 0.05, width = "260px"),
  sliderInput("te_sg", "Size of weekly shocks (sigma):",
              min = 1, max = 20, value = 8, step = 0.5, width = "260px"),
  sliderInput("te_n", "Ratings per client:",
              min = 4, max = 40, value = 21, step = 1, width = "260px"),
  actionButton("te_go", "New clients")
)

renderPlot({
  input$te_go
  wk <- 0:(input$te_n - 1)
  w  <- two_level_world(1000, 40, 1, 7, input$te_t1, 0, input$te_sg, weeks = wk)
  est <- ols_slopes(w$Y, wk)
  base_par()
  lim <- range(c(-2, 4, quantile(est, c(0.005, 0.995))))
  br  <- seq(floor(min(est, w$b1)) - 0.5, ceiling(max(est, w$b1)) + 0.5, by = 0.1)
  h1 <- hist(w$b1, breaks = br, plot = FALSE); h2 <- hist(est, breaks = br, plot = FALSE)
  plot(NA, xlim = lim, ylim = c(0, max(h1$density, h2$density) * 1.15), axes = FALSE,
       xlab = "change per week", ylab = "")
  abline(v = pretty(lim, 8), col = COL$grid)
  plot(h1, freq = FALSE, add = TRUE, col = alpha(COL$model, 0.45), border = NA)
  plot(h2, freq = FALSE, add = TRUE, col = alpha(COL$resid, 0.45), border = NA)
  axis(1, col = COL$grid, col.ticks = COL$muted)
  title(sprintf("%d ratings per client", input$te_n), adj = 0, line = 1.8,
        cex.main = 1.3, col.main = COL$ink)
  title(sprintf("SD of true slopes %.2f (blue)     SD of estimated slopes %.2f (orange)     theory %.2f",
                sd(w$b1), sd(est), sqrt(input$te_t1^2 + input$te_sg^2 / Sxx_n(input$te_n))),
        adj = 0, line = 0.5, cex.main = 0.98, col.main = COL$muted, font.main = 1)
}, height = 400)
```

$$\mathrm{SD}(\text{estimated slopes}) \;\approx\; \sqrt{\underbrace{\tau_1^2}_{\text{real differences}} + \underbrace{\sigma^2 / S_{xx}}_{\text{estimation noise}}}$$

::: {.question}
**Q10.** With 21 ratings, $\sigma = 8$ and $\tau_1 = 0.5$: how much wider is
the orange histogram than the blue one?

**Q11.** Now give each client only 5 ratings. What happens? Which clients end
up at the extremes of the orange histogram?

**Q12.** Set $\tau_1 = 0$: all clients truly improve at exactly the same rate.
What does a league table of estimated slopes show now?
:::

<details>
<summary>Show the answers</summary>

**Q10.** Only a little: about
`r f1(sqrt(0.25 + 64 / Sxx_n(21)), 2)` against 0.50. With 21 weekly ratings,
most of the spread in estimated slopes reflects real differences.

**Q11.** The orange histogram explodes, to an SD of about
`r f1(sqrt(0.25 + 64 / Sxx_n(5)), 1)`. With five ratings, the estimation
noise ($\sigma^2 / S_{xx} = 64 / 10$) swamps the real differences
($\tau_1^2 = 0.25$). The clients at the extremes are no longer the clients who
truly improve fastest or slowest; they are the ones whose few weeks happened
to fall luckily or unluckily.

**Q12.** A league table with winners and losers, even though every client
improves at exactly the same rate. All of the spread is noise. Ranking
clients by their estimated slopes, and especially calling the top and bottom
ones "best" and "worst", rewards and punishes luck, most of all for clients
with little data.

</details>

<div class="keybox">

**Three kinds of variability now.**
$\tau_0$ and $\tau_1$: how much clients **really** differ (between persons).
$\sigma$: how much a client's weeks scatter around their own line (within a
person).
And, from Lecture 1, the **estimation noise** in each client's slope,
$\sigma^2 / S_{xx}$, which is large for clients with few ratings.

To say where Albert stands, we need to separate the first from the last.
That is what the multilevel model does.

</div>

---

<div class="part">Part 2 · Albert's clinic</div>

# 4. Albert among 29 others

Albert's clinic has seen 30 clients in the same therapy, with the same weekly
questionnaire. `r sum(n_i == 21)` of them, including Albert, completed all 21
weeks; `r sum(n_i < 21)` stopped after between `r min(n_i)` and
`r max(n_i[n_i < 21])` ratings. (These clients are fictional, like Albert.)

```{r clinic-peek, echo=TRUE}
head(clinic)
table(ratings_per_client = n_i)
```

```{r fig-clinic, echo=FALSE}
inputPanel(
  selectInput("cl_who", "Highlight a client:", choices = label, selected = "Albert",
              width = "220px"),
  checkboxInput("cl_lines", "Show each client's own line", FALSE)
)

renderPlot({
  who <- match(input$cl_who, label)
  par(mfrow = c(5, 6)); base_par(mar = c(1.8, 2.2, 1.8, 0.4), mgp = c(1.6, 0.4, 0))
  for (j in ids) {
    x <- clinic[clinic$client == j, ]
    hl <- j == who
    plot(NA, xlim = c(0, 20), ylim = c(10, 95), axes = FALSE, xlab = "", ylab = "")
    abline(h = c(25, 50, 75), v = c(0, 10, 20), col = COL$grid)
    if (hl) rect(-2, 0, 22, 100, col = alpha(COL$resid, 0.08), border = NA)
    axis(1, at = c(0, 10, 20), col = COL$grid, cex.axis = 0.8)
    axis(2, at = c(25, 50, 75), col = COL$grid, cex.axis = 0.8)
    if (isTRUE(input$cl_lines) && nrow(x) > 1)
      segments(0, np[j, 1], max(x$week), np[j, 1] + np[j, 2] * max(x$week),
               col = if (hl) COL$resid else COL$model, lwd = 2)
    points(x$week, x$mood, pch = 19, cex = 0.8, col = if (hl) COL$resid else COL$ink)
    mtext(sprintf("%s (%d)", label[j], nrow(x)), side = 3, line = 0.2, adj = 0,
          cex = 0.75, font = if (hl) 2 else 1, col = if (hl) COL$resid else COL$ink)
  }
}, height = 640)
```

::: {.question}
**Q13.** Look at the clinic (the number in brackets is how many ratings each
client gave). Which clients seem to be improving fastest? Is anyone getting
worse? Write down your top three and your bottom one.
:::

---

# 5. Complete pooling: one line for everyone

The simplest way to use the whole clinic is to ignore who is who: pour all
`r nrow(clinic)` ratings into one regression, as if they came from one very
long Albert.

```{r cp, echo=TRUE}
m_cp <- lm(mood ~ week, data = clinic)
coef(summary(m_cp))
```

```{r fig-cp, echo=FALSE}
renderPlot({
  par(mfrow = c(1, 2)); base_par()
  panel(c(-0.6, 20.6), c(10, 95), xat = seq(0, 20, 5),
        main = "One line for everyone",
        sub = sprintf("mood = %.1f + %.2f x week", cp[1], cp[2]))
  points(jitter(clinic$week, 0.6), clinic$mood, pch = 19, cex = 0.7,
         col = alpha(COL$ink, 0.3))
  abline(m_cp, col = COL$model, lwd = 4)
  xa <- clinic$client == ID_A
  points(clinic$week[xa], clinic$mood[xa], pch = 19, cex = 1.2, col = COL$resid)
  legend("topleft", bty = "n", pch = 19, col = c(alpha(COL$ink, 0.4), COL$resid),
         legend = c("all clients", "Albert"), cex = 1)

  e  <- resid(m_cp)
  o  <- order(tapply(e, clinic$client, mean))
  boxplot(split(e, clinic$client)[o], horizontal = FALSE, axes = FALSE,
          col = ifelse(o == ID_A, alpha(COL$resid, 0.6), alpha(COL$model_lt, 0.6)),
          border = COL$muted, outline = FALSE, xlab = "clients (sorted)", ylab = "miss from the one line")
  abline(h = 0, col = COL$model, lwd = 2)
  axis(2, col = COL$grid, col.ticks = COL$muted)
  title("Misses, client by client", adj = 0, line = 1.8, cex.main = 1.3, col.main = COL$ink)
  title("each box = one client's misses; orange = Albert", adj = 0, line = 0.5,
        cex.main = 0.98, col.main = COL$muted, font.main = 1)
}, height = 420)
```

::: {.question}
**Q14.** What does this model tell Nora about Albert?

**Q15.** Look at the right panel. If the ratings really were `r nrow(clinic)`
independent weeks from one story, how should the boxes look? What do they
look like? And what does that mean for the standard error of
`r f1(se_cp, 3)`?
:::

<details>
<summary>Show the answers</summary>

**Q14.** Nothing about Albert in particular. It describes the typical course
in the clinic, about `r f1(cp[2], 2)` points a week, and treats Albert's 21
ratings as 21 of `r nrow(clinic)` interchangeable rows. It cannot say whether
Albert is doing better or worse than others, because it assumes nobody is
different.

**Q15.** In a single story with independent shocks, every box would straddle
zero. Instead, some clients' misses sit almost entirely above the line and
others almost entirely below. A client's weeks share that client's own start
and slope, so they are **not independent**: knowing one week's miss tells you
about the next. The model counts `r nrow(clinic)` independent pieces of
information when there are really far fewer, so its standard error of
`r f1(se_cp, 3)` is too small. (The multilevel model in Section 7 gives
`r f1(se_g[2], 3)` for the same quantity.) This is the independence problem
from Lecture 1, one level up.

</details>

::: {.tool}
<strong class="tag">The tool you just used</strong>

**Complete pooling**: one regression for all clients. It estimates the typical
course but assumes $\tau_0 = \tau_1 = 0$, and ignores that observations are
**clustered** within clients, which makes its standard errors too small.
:::

---

# 6. No pooling: a separate line for everyone

The opposite extreme: fit Lecture 1's model to every client separately. Each
client gets their own intercept and slope, estimated only from their own
ratings, as if no other client existed. (In one model:
`lm(mood ~ factor(client) * week)` gives the same lines.)

```{r np, echo=TRUE}
np <- t(sapply(split(clinic, clinic$client),
               function(x) coef(lm(mood ~ week, data = x))))
round(head(np), 2)       # row 1 is Albert: the same line as in Lecture 1
```

**The clinic's league table.** Every client's own slope, with its 95%
confidence interval.

```{r fig-league, echo=FALSE}
inputPanel(
  checkboxInput("lg_ci", "Show 95% confidence intervals", TRUE),
  checkboxInput("lg_n", "Colour by number of ratings", TRUE)
)

renderPlot({
  o <- order(np[, 2])
  lo <- np[, 2] - qt(0.975, pmax(n_i - 2, 1)) * np_se
  hi <- np[, 2] + qt(0.975, pmax(n_i - 2, 1)) * np_se
  cl <- if (isTRUE(input$lg_n)) ifelse(n_i < 21, COL$resid, COL$model) else rep(COL$model, 30)
  par(mfrow = c(1, 2)); base_par(mar = c(4.4, 6.5, 3.2, 1.2))
  xl <- if (isTRUE(input$lg_ci)) c(-8, 10) else c(-4, 5)
  plot(NA, xlim = xl, ylim = c(0.5, 30.5), axes = FALSE, xlab = "own slope (change per week)", ylab = "")
  abline(v = pretty(xl, 8), col = COL$grid); abline(v = 0, col = COL$muted)
  if (isTRUE(input$lg_ci)) segments(pmax(lo[o], xl[1]), 1:30, pmin(hi[o], xl[2]), 1:30, col = alpha(cl[o], 0.5), lwd = 2)
  points(np[o, 2], 1:30, pch = 19, cex = 1.3, col = cl[o])
  axis(1, col = COL$grid, col.ticks = COL$muted)
  axis(2, at = 1:30, labels = label[o], las = 1, cex.axis = 0.75, col = NA,
       font = 1)
  title("League table: own slopes", adj = 0, line = 1.8, cex.main = 1.3,
        col.main = COL$ink)
  title(if (isTRUE(input$lg_n)) "orange = fewer than 21 ratings" else "",
        adj = 0, line = 0.5, cex.main = 0.98, col.main = COL$muted, font.main = 1)

  base_par()
  plot(NA, xlim = c(3, 22), ylim = range(np[, 2]) + c(-0.5, 0.5), axes = FALSE,
       xlab = "number of ratings", ylab = "own slope")
  abline(h = pretty(np[, 2], 6), v = c(5, 10, 15, 20), col = COL$grid)
  abline(h = cp[2], col = COL$model, lwd = 2, lty = 2)
  points(jitter(n_i, 0.4), np[, 2], pch = 19, cex = 1.5, col = cl)
  text(n_i[c(ID_A, ID_J, ID_L)], np[c(ID_A, ID_J, ID_L), 2], label[c(ID_A, ID_J, ID_L)],
       pos = c(2, 4, 4), cex = 1, font = 2)
  axis(1, col = COL$grid, col.ticks = COL$muted); axis(2, col = COL$grid, col.ticks = COL$muted)
  title("Extremes = little data", adj = 0, line = 1.8, cex.main = 1.3,
        col.main = COL$ink)
  title("dashed = the one pooled line", adj = 0, line = 0.5, cex.main = 0.98,
        col.main = COL$muted, font.main = 1)
}, height = 560)
```

::: {.question}
**Q16.** By this table, Jonas is the clinic's most improved client and Lena
is getting worse fast. Would you give Jonas an award, or call Lena's family?
Compare with your picks from Q13.

**Q17.** What do the clients at the very top and bottom of the table have in
common?

**Q18.** No pooling estimates each client as if the other 29 did not exist.
What useful information does it throw away?
:::

<details>
<summary>Show the answers</summary>

**Q16.** Neither, yet. Jonas's `r f1(np[ID_J, 2], 1)` points a week and Lena's
`r f1(np[ID_L, 2], 1)` each come from only `r n_i[ID_J]` ratings; their
confidence intervals run from about `r round(np[ID_J, 2] - qt(0.975, n_i[ID_J] - 2) * np_se[ID_J])`
to `r round(np[ID_J, 2] + qt(0.975, n_i[ID_J] - 2) * np_se[ID_J])` and from
`r round(np[ID_L, 2] - qt(0.975, n_i[ID_L] - 2) * np_se[ID_L])` to
`r round(np[ID_L, 2] + qt(0.975, n_i[ID_L] - 2) * np_se[ID_L])` points a week.
Five weeks with one lucky or unlucky rating at the end is enough to produce
such slopes. This is Section 3, in real data.

**Q17.** They are almost all clients who dropped out early. The funnel on the
right shows it: with few ratings, own slopes scatter wildly; with 21 they
cluster. The extremes of a no-pooling league table are dominated by the
clients we know least about.

**Q18.** Everything the clinic has learned about clients in general: that
clients here typically gain around `r f1(cp[2], 1)` points a week, and that
rates of change far beyond ±3 almost never happen among the clients with
plenty of data. Faced with a client who has 5 ratings, a sensible clinician
would combine those 5 ratings with that experience. No pooling doesn't.

</details>

::: {.tool}
<strong class="tag">The tool you just used</strong>

**No pooling**: a separate regression (fixed intercept and slope) for each
client, equivalently a **fixed-effects** model with a dummy for every client.
It lets clients differ, but learns nothing from one client about another, and
its estimates for clients with little data are very noisy.
:::

---

# 7. The multilevel model: both levels at once

Complete pooling says clients are all the same. No pooling says they have
nothing in common. The two-level story from Part 1 says something in between:
**clients differ, but they come from the same population.** Fitting that
story is what a multilevel (mixed-effects) model does:

```{r re, echo=TRUE}
m_re <- lmer(mood ~ week + (week | client), data = clinic)
summary(m_re)
```

`(week | client)` means: every client gets their own intercept and their own
slope for `week`, and those are drawn from a common population. The output
estimates the dials of the two-level story:

| In the output | Value | Dial | In words |
|---|---|---|---|
| Fixed effect `(Intercept)` | `r f1(g_hat[1])` | $\gamma_{00}$ | the typical client's mood at intake |
| Fixed effect `week` | `r f1(g_hat[2], 2)` (SE `r f1(se_g[2], 3)`) | $\gamma_{10}$ | the typical client's change per week |
| Random effect SD `(Intercept)` | `r f1(tau0_h)` | $\tau_0$ | how much clients differ in where they start |
| Random effect SD `week` | `r f1(tau1_h, 2)` | $\tau_1$ | how much clients differ in how fast they change |
| `Corr` | `r f1(rho_h, 2)` | $\rho$ | clients who start lower tend to improve faster |
| Residual SD | `r f1(sig_h)` | $\sigma$ | the size of life's weekly shocks |

Now Nora's question has an answer. The clinic's population of rates of
change is roughly a bell centred on `r f1(g_hat[2], 2)` with SD
`r f1(tau1_h, 2)`. Albert's own rate, as the multilevel model estimates it, is
**`r f1(pp[ID_A, 2], 2)`**, faster than about **`r pct(pct_A_pp)`%** of
clients.

**Is Albert on track?** The band shows where the clinic's clients' expected
mood usually lies at each week, the way routine outcome monitoring draws
expected recovery curves.

```{r fig-track, echo=FALSE}
inputPanel(
  selectInput("tr_who", "Client:", choices = label, selected = "Albert", width = "200px"),
  checkboxInput("tr_shock", "Band for single weeks (add the weekly shocks)", FALSE)
)

renderPlot({
  j <- match(input$tr_who, label); x <- clinic[clinic$client == j, ]
  t <- seq(0, 20, 0.1)
  mu <- g_hat[1] + g_hat[2] * t
  v  <- tau0_h^2 + t^2 * tau1_h^2 + 2 * t * rho_h * tau0_h * tau1_h +
        if (isTRUE(input$tr_shock)) sig_h^2 else 0
  base_par()
  panel(c(-0.6, 20.6), c(0, 100), xat = seq(0, 20, 5),
        main = sprintf("%s against the clinic's expected course", label[j]),
        sub = if (isTRUE(input$tr_shock)) "bands: where 50% and 90% of single weekly ratings fall"
              else "bands: where 50% and 90% of clients' expected mood lies")
  polygon(c(t, rev(t)), c(mu - 1.645 * sqrt(v), rev(mu + 1.645 * sqrt(v))),
          col = alpha(COL$model_lt, 0.35), border = NA)
  polygon(c(t, rev(t)), c(mu - 0.674 * sqrt(v), rev(mu + 0.674 * sqrt(v))),
          col = alpha(COL$model_lt, 0.55), border = NA)
  lines(t, mu, col = COL$model, lwd = 3)
  segments(0, pp[j, 1], 20, pp[j, 1] + 20 * pp[j, 2], col = COL$resid, lwd = 3)
  points(x$week, x$mood, pch = 19, cex = 1.5, col = COL$ink)
  legend("bottomright", bty = "n", lwd = c(3, 3, NA), pch = c(NA, NA, 19),
         col = c(COL$model, COL$resid, COL$ink), cex = 1,
         legend = c("typical client", sprintf("%s (multilevel estimate)", label[j]),
                    "ratings"))
}, height = 420)
```

::: {.question}
**Q19.** Is Albert on track? From about when does he move above the typical
client's course?

**Q20.** Tick "band for single weeks". Why is the band so much wider, and
which band should Nora use to decide whether *one* bad week is alarming?

**Q21.** The typical change per week has SE `r f1(se_g[2], 3)` here, against
`r f1(se_cp, 3)` under complete pooling. Why is the multilevel SE larger, and
which one is right?
:::

<details>
<summary>Show the answers</summary>

**Q19.** Albert starts a little below the typical client (his estimated
intake level is `r f1(pp[ID_A, 1])` against `r f1(g_hat[1])`) but climbs
faster, crosses the typical course within the first couple of weeks, and by week 20
is in the upper part of the band, near its 90% edge. On this questionnaire, in this clinic, therapy is going
well for Albert.

**Q20.** The narrow band is about clients' *expected* mood, their lines; the
wide one adds the weekly shocks, so it describes single ratings. It is the
same distinction as confidence versus prediction bands in Lecture 1. To judge
one bad week, Nora needs the wide band; to judge a client's overall course,
the narrow one.

**Q21.** The multilevel model knows the ratings come from 30 clients who
genuinely differ. The typical rate is learned from 30 clients' rates, not from
`r nrow(clinic)` independent weeks, so there is more uncertainty about it. The
multilevel SE is the honest one; complete pooling's is too small (Q15).

</details>

::: {.tool}
<strong class="tag">The tool you just used</strong>

The **linear mixed-effects (multilevel) model** with **random intercepts and
random slopes**, `lmer(y ~ time + (time | person))`: **fixed effects**
$\gamma_{00}, \gamma_{10}$ describe the typical person; **variance
components** $\tau_0, \tau_1, \rho$ describe how persons differ; $\sigma$ is
the within-person residual SD.
:::

---

# 8. Partial pooling: why Jonas moves and Albert hardly does

The multilevel model's line for each client is not their no-pooling line.
Each client's estimate is **pulled towards the typical client**, some a lot,
some hardly at all.

```{r fig-shrink, echo=FALSE}
inputPanel(
  radioButtons("sh_view", "Show:", inline = TRUE,
               choices = c("slopes side by side" = "side",
                           "start and slope together" = "plane"), selected = "side")
)

renderPlot({
  cl <- ifelse(n_i < 21, COL$resid, COL$model)
  key <- c(ID_A, ID_J, ID_L)
  base_par()
  if (input$sh_view == "side") {
    plot(NA, xlim = c(0.6, 2.9), ylim = range(np[, 2]) + c(-0.3, 0.3), axes = FALSE,
         xlab = "", ylab = "change per week")
    abline(h = pretty(np[, 2], 8), col = COL$grid)
    abline(h = g_hat[2], col = COL$muted, lty = 2, lwd = 2)
    segments(1, np[, 2], 2, pp[, 2], col = alpha(cl, 0.6), lwd = ifelse(ids %in% key, 3, 1.5))
    points(rep(1, 30), np[, 2], pch = 21, bg = "white", col = cl, cex = 1.6, lwd = 2)
    points(rep(2, 30), pp[, 2], pch = 19, col = cl, cex = 1.6)
    text(2.05, pp[key, 2], sprintf("%s (%d ratings)", label[key], n_i[key]), pos = 4,
         font = 2, cex = 1.05)
    axis(2, col = COL$grid, col.ticks = COL$muted)
    axis(1, at = 1:2, labels = c("no pooling\n(own data only)", "partial pooling\n(multilevel)"),
         col = NA, cex.axis = 1.1, padj = 0.5)
    title("Every client pulled towards the typical client", adj = 0, line = 1.8,
          cex.main = 1.3, col.main = COL$ink)
    title("dashed = typical change per week;  orange = dropped out early",
          adj = 0, line = 0.5, cex.main = 0.98, col.main = COL$muted, font.main = 1)
  } else {
    xl <- range(np[, 1], pp[, 1]) + c(-3, 3); yl <- range(np[, 2], pp[, 2]) + c(-0.3, 0.3)
    plot(NA, xlim = xl, ylim = yl, axes = FALSE, xlab = "starting level (intercept)",
         ylab = "change per week (slope)")
    abline(h = pretty(yl, 8), v = pretty(xl, 8), col = COL$grid)
    V <- matrix(c(tau0_h^2, rho_h * tau0_h * tau1_h, rho_h * tau0_h * tau1_h, tau1_h^2), 2)
    th <- seq(0, 2 * pi, length.out = 200); ev <- eigen(V)
    for (r in c(1, 2)) {
      E <- t(g_hat + r * ev$vectors %*% diag(sqrt(ev$values)) %*% rbind(cos(th), sin(th)))
      lines(E, col = COL$muted, lty = r, lwd = 1.5)
    }
    arrows(np[, 1], np[, 2], pp[, 1], pp[, 2], length = 0.08, col = alpha(cl, 0.7),
           lwd = ifelse(ids %in% key, 3, 1.5))
    points(np[, 1], np[, 2], pch = 21, bg = "white", col = cl, cex = 1.4, lwd = 2)
    points(pp[, 1], pp[, 2], pch = 19, col = cl, cex = 1.4)
    points(g_hat[1], g_hat[2], pch = 4, cex = 2.5, lwd = 4, col = COL$ink)
    text(pp[key, 1], pp[key, 2], label[key], pos = 4, font = 2, cex = 1.05)
    axis(1, col = COL$grid, col.ticks = COL$muted); axis(2, col = COL$grid, col.ticks = COL$muted)
    title("From own estimate (open) to multilevel estimate (filled)", adj = 0, line = 1.8,
          cex.main = 1.3, col.main = COL$ink)
    title("cross = typical client; ellipses = where 1 and 2 SDs of clients lie",
          adj = 0, line = 0.5, cex.main = 0.98, col.main = COL$muted, font.main = 1)
  }
}, height = 520)
```

Why do some estimates move more than others? For the slope, the multilevel
estimate is approximately a **weighted average** of the client's own estimate
and the typical client's:

$$\hat\beta_{1i}^{\text{pooled}} \;\approx\; w_i\, \hat\beta_{1i}^{\text{own}} + (1 - w_i)\, \hat\gamma_{10},
\qquad w_i = \frac{\tau_1^2}{\tau_1^2 + \sigma^2 / S_{xx,i}}$$

The weight compares **how much clients really differ** ($\tau_1^2$) with
**how noisy this client's own estimate is** ($\sigma^2 / S_{xx,i}$). Lots of
data: the own estimate is precise, $w_i$ is close to 1, and the client keeps
their own slope. Little data: $w_i$ is close to 0, and the client is pulled
towards the typical client. (The real computation also uses the intercept and
its correlation with the slope, so the formula is approximate.)

```{r fig-weight, echo=FALSE}
inputPanel(
  sliderInput("wt_n", "Ratings for this client:", min = 3, max = 40, value = 5,
              step = 1, width = "240px"),
  sliderInput("wt_t1", "How much clients differ (tau1):", min = 0.05, max = 2,
              value = round(tau1_h, 2), step = 0.05, width = "240px"),
  sliderInput("wt_own", "This client's own slope:", min = -4, max = 5,
              value = round(np[ID_J, 2], 1), step = 0.1, width = "240px")
)

renderPlot({
  ns <- 3:40
  ws <- w_n(ns, input$wt_t1, sig_h)
  w  <- w_n(input$wt_n, input$wt_t1, sig_h)
  est <- w * input$wt_own + (1 - w) * g_hat[2]
  par(mfrow = c(1, 2)); base_par()
  plot(NA, xlim = c(3, 40), ylim = c(0, 1), axes = FALSE, xlab = "number of weekly ratings",
       ylab = "weight on the client's own slope")
  abline(h = seq(0, 1, 0.25), v = seq(5, 40, 5), col = COL$grid)
  abline(h = 0.5, col = COL$muted, lty = 2)
  lines(ns, ws, col = COL$model, lwd = 4)
  points(input$wt_n, w, pch = 19, cex = 2.2, col = COL$resid)
  axis(1, col = COL$grid, col.ticks = COL$muted); axis(2, col = COL$grid, col.ticks = COL$muted)
  title("Weight on own data", adj = 0, line = 1.8, cex.main = 1.3,
        col.main = COL$ink)
  title(sprintf("with %d ratings: weight %.2f", input$wt_n, w), adj = 0, line = 0.5,
        cex.main = 0.98, col.main = COL$muted, font.main = 1)

  xl <- c(-5, 6); plot(NA, xlim = xl, ylim = c(0, 3), axes = FALSE,
                              xlab = "change per week", ylab = "")
  abline(v = pretty(xl, 9), col = COL$grid)
  arrows(input$wt_own, 2, est, 1, length = 0.15, lwd = 3, col = COL$resid)
  points(input$wt_own, 2, pch = 21, bg = "white", col = COL$resid, cex = 3, lwd = 3)
  points(est, 1, pch = 19, col = COL$resid, cex = 3)
  abline(v = g_hat[2], col = COL$model, lwd = 3, lty = 2)
  text(input$wt_own, 2.35, "own estimate", cex = 1.1)
  text(est, 0.6, sprintf("pooled: %.2f", est), cex = 1.1, font = 2)
  text(g_hat[2], 2.85, "typical client", col = COL$model, cex = 1.1)
  axis(1, col = COL$grid, col.ticks = COL$muted)
  title("Pulled towards typical", adj = 0, line = 1.8,
        cex.main = 1.3, col.main = COL$ink)
}, height = 400)
```

::: {.question}
**Q22.** In this clinic, how many weekly ratings does a client need before
their own data get more than half the weight?

**Q23.** Albert has 21 ratings, Jonas 5. Read off each one's weight. Why does
Jonas's estimate move from `r f1(np[ID_J, 2], 1)` to `r f1(pp[ID_J, 2], 2)`,
while Albert's only moves from `r f1(np[ID_A, 2], 2)` to
`r f1(pp[ID_A, 2], 2)`?

**Q24.** Set "how much clients differ" to its smallest value, then to its
largest. What happens to everyone's weight, and why does that make sense?
:::

<details>
<summary>Show the answers</summary>

**Q22.** About **`r n_half`** weekly ratings. Below that, the clinic's
experience counts for more than the client's own data; above it, the client's
own data dominate.

**Q23.** Albert's own slope is precise (21 ratings, weight about
`r f1(w_n(21), 2)`), so the model mostly believes it and pulls it only
slightly towards `r f1(g_hat[2], 2)`. Jonas's own slope is extremely noisy
(5 ratings, weight about `r f1(w_n(5), 2)`), so the model mostly relies on
what clients in this clinic usually do. It's what an experienced clinician
does informally: "five good weeks is encouraging, but I've seen that before."
(The model's value, `r f1(pp[ID_J, 2], 2)`, is a bit above the formula's
`r f1(g_hat[2] + w_n(5) * (np[ID_J, 2] - g_hat[2]), 2)` because the model also
uses Jonas's low starting level: in this clinic, clients who start low tend to
improve faster, $\rho < 0$.)

**Q24.** If clients hardly differ ($\tau_1$ small), there is little to learn
from any single client, so everyone is pulled almost all the way to the
typical client: close to complete pooling. If clients differ enormously
($\tau_1$ large), knowing the typical client tells you little about any one
client, so everyone keeps their own estimate: close to no pooling. Partial
pooling is not a fixed compromise; the data decide, through $\tau_1$, how
much to pool.

</details>

### Is shrinking fair? A check against the truth

Pulling Jonas down and Lena up can look like cheating: the model overrides
what their own ratings say. Is it better? Usually we can't check. Here we
can: **Albert's 29 fellow clients were simulated**, so for them (not for
Albert) we know their true rates of change.

```{r fig-truth, echo=FALSE}
renderPlot({
  cl <- ifelse(n_i[-1] < 21, COL$resid, COL$model)
  par(mfrow = c(1, 2)); base_par()
  lim <- range(c(np[-1, 2], pp[-1, 2], TRUE_B1_OTHERS)) + c(-0.3, 0.3)
  for (k in 1:2) {
    est <- if (k == 1) np[-1, 2] else pp[-1, 2]
    rm  <- sqrt(mean((est - TRUE_B1_OTHERS)^2))
    plot(NA, xlim = range(TRUE_B1_OTHERS) + c(-0.3, 0.3), ylim = lim, axes = FALSE,
         xlab = "true change per week", ylab = "estimated change per week")
    abline(h = pretty(lim, 6), v = pretty(TRUE_B1_OTHERS, 6), col = COL$grid)
    abline(0, 1, col = COL$muted, lwd = 2)
    segments(TRUE_B1_OTHERS, TRUE_B1_OTHERS, TRUE_B1_OTHERS, est, col = alpha(cl, 0.5), lwd = 2)
    points(TRUE_B1_OTHERS, est, pch = if (k == 1) 21 else 19, bg = "white", col = cl,
           cex = 1.6, lwd = 2)
    kk <- c(ID_J, ID_L) - 1
    text(TRUE_B1_OTHERS[kk], est[kk], label[kk + 1], pos = 4, font = 2)
    axis(1, col = COL$grid, col.ticks = COL$muted); axis(2, col = COL$grid, col.ticks = COL$muted)
    title(if (k == 1) "No pooling" else "Partial pooling", adj = 0, line = 1.8,
          cex.main = 1.3, col.main = COL$ink)
    title(sprintf("typical error (RMSE) = %.2f points per week", rm), adj = 0, line = 0.5,
          cex.main = 0.98, col.main = COL$muted, font.main = 1)
  }
}, height = 420)
```

Jonas's true rate of change is **`r f1(TRUE_B1_OTHERS[ID_J - 1], 2)`**
(no pooling said `r f1(np[ID_J, 2], 2)`, partial pooling
`r f1(pp[ID_J, 2], 2)`). Lena's is **`r f1(TRUE_B1_OTHERS[ID_L - 1], 2)`**:
she was improving all along (no pooling said `r f1(np[ID_L, 2], 2)`, partial
pooling `r f1(pp[ID_L, 2], 2)`). Across the 29 clients, partial pooling's
typical error is `r f1(rmse_pp, 2)` points a week against no pooling's
`r f1(rmse_np, 2)`.

**Is that luck?** Open new simulated clinics and compare.

```{r fig-sim-rmse, echo=FALSE}
inputPanel(
  sliderInput("rm_t1", "How much clients differ (tau1):", min = 0.1, max = 1.5,
              value = 0.5, step = 0.05, width = "240px"),
  sliderInput("rm_drop", "Clients who drop out early:", min = 0, max = 25,
              value = 9, step = 1, width = "240px"),
  actionButton("rm_go", "Open a new clinic")
)

renderPlot({
  input$rm_go
  J <- 30
  w <- two_level_world(J, 40, 1, 7, input$rm_t1, -0.3, 8)
  nn <- rep(21, J); if (input$rm_drop > 0) nn[seq_len(input$rm_drop)] <- sample(4:10, input$rm_drop, TRUE)
  d <- do.call(rbind, lapply(1:J, function(j)
    data.frame(client = j, week = 0:(nn[j] - 1), mood = w$Y[1:nn[j], j])))
  own <- sapply(split(d, d$client), function(x) coef(lm(mood ~ week, x))[2])
  m <- suppressWarnings(suppressMessages(lmer(mood ~ week + (week | client), data = d)))
  pool <- coef(m)$client[, 2]
  r1 <- sqrt(mean((own - w$b1)^2)); r2 <- sqrt(mean((pool - w$b1)^2))
  base_par()
  lim <- range(c(own, pool, w$b1))
  plot(NA, xlim = range(w$b1) + c(-0.2, 0.2), ylim = lim, axes = FALSE,
       xlab = "true change per week", ylab = "estimate")
  abline(h = pretty(lim, 6), v = pretty(w$b1, 6), col = COL$grid)
  abline(0, 1, col = COL$muted, lwd = 2)
  points(w$b1, own, pch = 21, bg = "white", col = COL$resid, cex = 1.5, lwd = 2)
  points(w$b1, pool, pch = 19, col = COL$model, cex = 1.5)
  legend("topleft", bty = "n", pch = c(21, 19), col = c(COL$resid, COL$model), pt.lwd = 2,
         legend = c(sprintf("no pooling: RMSE %.2f", r1), sprintf("partial pooling: RMSE %.2f", r2)),
         cex = 1.1)
  axis(1, col = COL$grid, col.ticks = COL$muted); axis(2, col = COL$grid, col.ticks = COL$muted)
  title("A new simulated clinic", adj = 0, line = 1.8, cex.main = 1.3, col.main = COL$ink)
}, height = 420)
```

::: {.question}
**Q25.** Open several new clinics. Does partial pooling win every time? When
is its advantage largest, and when does it nearly vanish?

**Q26.** Partial pooling is *biased* for individual clients: it pulls genuine
outliers towards the middle. When could that be clinically dangerous?
:::

<details>
<summary>Show the answers</summary>

**Q25.** Almost always, and by most when many clients have few ratings and
clients differ little (small $\tau_1$), because then own estimates are mostly
noise. With few dropouts and large $\tau_1$ the two methods nearly agree:
there is little to pool. The price of partial pooling is a little bias for
each client; the gain is a large reduction in noise. On average, that trade
wins.

**Q26.** A client who really is deteriorating fast, but has given only a few
ratings, will be pulled towards "improving like everyone else". If a clinic
uses shrunken estimates to decide who needs extra attention, genuine early
warning signs can be softened. Early-warning systems therefore often look at
individual ratings too, not only at pooled trajectories. A second danger: if
clients drop out *because* therapy isn't working for them, the population
itself is estimated from a biased sample, and everyone is pulled towards a
too-optimistic "typical client".

</details>

::: {.tool}
<strong class="tag">The tools you just used</strong>

**Partial pooling** (also called **shrinkage** or **borrowing strength**):
each person's estimate combines their own data with the population estimate,
weighted by their precision. The resulting person-specific estimates are the
model's **conditional modes** (often called **BLUPs**), `coef(m)` or
`ranef(m)` in `lme4`. They are biased towards the mean, but more accurate on
average than no-pooling estimates.
:::

---

# 9. What Lecture 2 lets us say about Albert

::: {.question}
**Q27.** Write two or three sentences for Nora: is therapy going well for
Albert, compared with other clients in the clinic? Include one caveat.
:::

<details>
<summary>Show one version</summary>

<div class="report">

In this clinic, clients' mood typically rises by about `r f1(g_hat[2], 2)`
points a week (95% CI roughly `r f1(g_hat[2] - 1.96 * se_g[2], 1)` to
`r f1(g_hat[2] + 1.96 * se_g[2], 1)`), and clients differ in their rate of
change with an SD of about `r f1(tau1_h, 2)`. Albert's estimated rate,
borrowing strength from the whole clinic, is about `r f1(pp[ID_A, 2], 1)`
points a week, faster than roughly `r pct(pct_A_pp)`% of clients. He started a
little below the typical client and is now well above the expected course.

*Caveat:* the comparison assumes the clinic's clients, including those who
dropped out, are a fair picture of clients like Albert; the percentage is
read off a bell-shaped population of change rates (Section 2); and it still
assumes that each client's weeks follow a straight line with memoryless
shocks.

</div>

</details>

<div class="keybox">

**The model of Lecture 2**

$$\text{Mood}_{it} = \beta_{0i} + \beta_{1i}\,t + \varepsilon_{it}, \qquad
\varepsilon_{it} \sim \mathcal{N}(0, \sigma)$$

$$\begin{pmatrix}\beta_{0i}\\ \beta_{1i}\end{pmatrix} \sim
\mathcal{N}\!\left(\begin{pmatrix}\gamma_{00}\\ \gamma_{10}\end{pmatrix},
\begin{pmatrix}\tau_0^2 & \rho\tau_0\tau_1\\ \rho\tau_0\tau_1 & \tau_1^2\end{pmatrix}\right)$$

| Model | What it assumes about clients | In R |
|---|---|---|
| Complete pooling | all the same ($\tau_0 = \tau_1 = 0$) | `lm()` on all ratings |
| No pooling | unrelated to each other ($\tau \to \infty$) | `lm()` per client |
| Partial pooling | different, but from one population ($\tau$ estimated) | `lmer()` with random intercepts and slopes |

In code: `lm(mood ~ week)`, `lm(mood ~ factor(client) * week)` and
`lmer(mood ~ week + (week | client))`.

Lecture 1 asked how Albert changes. Lecture 2 adds: **how does Albert's
change compare with others', and what can others tell us about Albert?** It
adds one assumption, that clients' dials come from a single bell-shaped
population, which is a claim about how clients come to differ. Both
lectures still assume each client changes along a straight line with
memoryless weekly shocks. Later lectures keep the two-level idea and change
the within-person story, so we can ask, for instance, whether Albert's mood is
more "sticky" from week to week than other clients'.

</div>

---

# 10. Exercises

**E1. Do clients really differ in how fast they change?**
Fit a model in which clients differ only in where they start:
`m1 <- lmer(mood ~ week + (1 | client), data = clinic)`. What does this model
claim about therapy? Compare it with `m_re` using `anova(m1, m_re)`. What do
you conclude?

**E2. Where you put zero, again.**
Refit with week centred at mid-therapy:
`lmer(mood ~ I(week - 10) + (I(week - 10) | client), data = clinic)`. What
happens to $\tau_0$ and to the correlation $\rho$? Relate this to Section 11
of Lecture 1.

**E3. How much of mood is "who you are"?**
Fit `m0 <- lmer(mood ~ 1 + (1 | client), data = clinic)` and compute the
**intraclass correlation**, $\tau_0^2 / (\tau_0^2 + \sigma^2)$. In words, what
share of the variation in weekly ratings lies between clients rather than
within them?

**E4. When dropping out is not random.**
Adapt the simulation in Section 8 so that clients whose true slope is below
0.5 are more likely to drop out early. Fit the multilevel model to many such
clinics. What happens to the estimated typical change per week, and why does
it matter for Albert's comparison?

**E5. A new client.**
A new client, Sam, has given three ratings: 38, 45 and 41. Estimate Sam's
rate of change with no pooling, then add Sam to `clinic` and refit `m_re`.
Compare Sam's multilevel slope with his own slope, and explain the difference
using the weight formula.

**E6. Can the model see the two kinds of clients?**
Simulate a clinic of 30 clients with 21 weekly ratings each, where 85% are
responders (true slope drawn from $\mathcal{N}(1.5, 0.3)$) and 15% are
non-responders ($\mathcal{N}(-0.1, 0.3)$); use the intercepts and $\sigma$
of Section 1. Fit `lmer(mood ~ week + (week | client))`. Compare
$\hat\gamma_{10}$ and $\hat\tau_1$ with the mean and SD of the true slopes.
Then compare a histogram of the model's client slopes (`coef(m)$client$week`)
with one of the true slopes. Repeat for a few clinics, then with only 6
ratings per client. Why is looking at the model's own client estimates a weak
check of the bell?

<details>
<summary>Hints</summary>

**E1.** The random-intercepts model says all clients improve at the same rate.
The likelihood-ratio test compares the two models; with 30 clients the random
slopes should clearly improve the fit. Look at the $\tau_1$ estimate as well
as the p-value.

**E3.** In this clinic the ICC is about
`r f1({v0 <- as.data.frame(VarCorr(lmer(mood ~ 1 + (1 | client), data = clinic)))$vcov; v0[1] / sum(v0)}, 2)`.

**E5.** Sam's own slope comes from three ratings, so its weight is tiny (about
`r f1(w_n(3), 2)`); his multilevel slope will be close to the typical
client's.

**E6.** The typical change and its SD are usually close to the mean and SD of
the true slopes: those estimates hold up well when the bell is wrong. With
21 ratings, the model's client slopes often still show the two kinds, but
blurred. With 6 ratings the gap largely disappears: estimation noise fills it
in, and every client is pulled strongly towards the single centre the bell
assumes, so the estimates take on the bell's shape. To see the two kinds, the
story has to allow for them, for example with a mixture model.

</details>

---

<p style="color:#898781; font-size:0.9em; margin-top:3em;">
Albert, Nora and the clinic's clients are fictional. The clinic's other
clients were simulated from the two-level story of Section 1.
Rendered with R, Shiny and lme4.
</p>
